---
title: "Areal CAR models with stLMM"
subtitle: "County graphs, island handling, recovery, and holdout prediction"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    code-overflow: wrap
    html-math-method: katex
execute:
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| code-fold: true
#| code-summary: "Show preliminaries"
library(stLMM)
library(ggplot2)
library(sf)

source(file.path(dirname(knitr::current_input(dir = TRUE)), "utils.R"))

set.seed(1)
```

```{r plot-theme, include=FALSE}
theme_set(theme_bw(base_size = 12))
```

## Overview

This vignette introduces areal spatial models using `car()` to fit conditional autoregressive (CAR) terms. To motivate this example, we use the bundled `stunitco` county polygon layer, extract Washington counties, build a county adjacency graph, simulate a Gaussian response from a proper CAR model, fit the model with a few county responses held out, recover the county random effects, and predict the held-out responses.

The key difference from point-referenced Gaussian process models is that the latent process is indexed by areal units (counties in this case) rather than coordinates. The graph defines which areas are neighbors, and the CAR precision matrix defines the dependence among area random effects.

## Model

For area $i$, the model is

$$
y_i = \beta_0 + x_i\beta_1 + w_i + \epsilon_i,\qquad
\epsilon_i \sim N(0,\tau^2).
$$

The default areal process is specified through the proper CAR precision matrix

$$
\mathbf{Q} = \sigma^{-2}(\mathbf{D} - \rho\mathbf{G}),
$$

where $\mathbf{G}$ is the binary adjacency matrix, $\mathbf{D}$ is diagonal with entries equal to the number of neighbors for each area, $\sigma^2$ controls the process variance, and $\rho$ controls spatial association. The graph itself stores binary adjacency, but this degree-scaled precision is equivalent to a CAR conditional mean that averages neighboring effects. See @BanerjeeCarlinGelfand2014 for background on CAR models and hierarchical spatial modeling for areal data.

The same `car_graph()` object can also be used with `car_model = "leroux"`, which gives the Leroux CAR precision $\sigma^{-2}\{(1-\rho)\mathbf{I}+\rho(\mathbf{D}-\mathbf{G})\}$ [@LerouxLeiBreslow1999]. This alternative is useful when the model should be able to move between nearly iid area effects and strong spatial smoothing using the same `rho` parameter. In this introductory example we use the default proper CAR model.

## County graph

The `stunitco` object contains county polygons for the conterminous United States. Washington has one county that is not connected by polygon boundaries to the rest of the state graph, so the default graph-building rule reports the issue in the error below. For this vignette, we make the island rule explicit and add nearest-neighbor bridge edges.

```{r wa-data}
data(stunitco, package = "stLMM")

wa_counties <- stunitco[stunitco$STATECD == 53, ]
```

```{r wa-graph-error}
#| error: true
car_graph(wa_counties, id = "COUNTYFIPS")
```

Now, let's correct the island issue using the nearest-neighbor bridge rule in the code below. Here, setting `island_k = 4` adds four nearest-neighbor bridge edges from the disconnected component.

```{r wa-graph}
g <- car_graph(wa_counties, id = "COUNTYFIPS", island = "nearest", island_k = 4)

g$island_added_edges
```

The graph object stores county IDs, binary adjacency, node degrees, and any bridge edges added by the island rule. The graph is not stored as a row-standardized matrix; the neighbor averaging enters through the degree-scaled CAR precision.

```{r wa-edge-graph, fig.width=6.2, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show edge-map code"
coord_dat <- data.frame(
  area = as.character(wa_counties$COUNTYFIPS),
  sf::st_coordinates(
    sf::st_point_on_surface(sf::st_geometry(wa_counties))
  )
)

edge_index <- which(as.matrix(g$adjacency) != 0, arr.ind = TRUE)
edge_index <- edge_index[edge_index[, "row"] < edge_index[, "col"], , drop = FALSE]
edge_dat <- data.frame(
  from = g$ids[edge_index[, "row"]],
  to = g$ids[edge_index[, "col"]]
)
edge_dat$key <- paste(
  pmin(edge_dat$from, edge_dat$to),
  pmax(edge_dat$from, edge_dat$to),
  sep = "--"
)
edge_dat$x <- coord_dat$X[match(edge_dat$from, coord_dat$area)]
edge_dat$y <- coord_dat$Y[match(edge_dat$from, coord_dat$area)]
edge_dat$xend <- coord_dat$X[match(edge_dat$to, coord_dat$area)]
edge_dat$yend <- coord_dat$Y[match(edge_dat$to, coord_dat$area)]

island_key <- paste(
  pmin(g$island_added_edges$from, g$island_added_edges$to),
  pmax(g$island_added_edges$from, g$island_added_edges$to),
  sep = "--"
)
edge_dat$island <- edge_dat$key %in% island_key

ggplot(wa_counties) +
  geom_sf(
    fill = "grey95",
    color = "white",
    linewidth = 0.15
  ) +
  geom_segment(
    data = edge_dat,
    aes(x = x, y = y, xend = xend, yend = yend, color = island, linewidth = island)
  ) +
  geom_point(
    data = coord_dat,
    aes(X, Y),
    color = stlmm_color("primary"),
    size = 1.4
  ) +
  scale_color_manual(
    values = c("FALSE" = "grey45", "TRUE" = stlmm_color("secondary")),
    guide = "none"
  ) +
  scale_linewidth_manual(
    values = c("FALSE" = 0.25, "TRUE" = 1),
    guide = "none"
  ) +
  labs(x = "longitude", y = "latitude")
```

## Data

We construct a simple county-level covariate from the east-west position of each polygon. This is only for simulation; it gives the fixed-effect part of the model a visible spatial trend. Following the model statement, the CAR dependence is parameterized through the precision matrix $\mathbf{Q}$. However, in the simulation code below, `rmvnorm()` draws from a covariance matrix, so we compute $\mathbf{\Sigma} = \mathbf{Q}^{-1}$ before drawing the county effects.

```{r simulate-data}
centroid_x <- sf::st_coordinates(sf::st_centroid(sf::st_geometry(wa_counties)))[, "X"]
centroid_x_mean <- mean(centroid_x)
centroid_x_sd <- stats::sd(centroid_x)

dat <- data.frame(
  area = as.character(wa_counties$COUNTYFIPS),
  county = wa_counties$COUNTYNM,
  x = (centroid_x - centroid_x_mean) / centroid_x_sd
)

beta <- c(0.5, -1)
sigma_sq <- 1
rho <- 0.75
tau_sq <- 0.1

Q <- car_prec(g, sigma_sq = sigma_sq, rho = rho)
Sigma <- solve(Q)

w <- rmvnorm(mean = rep(0, nrow(Q)), Sigma = Sigma)
names(w) <- rownames(Q)

dat$w_true <- w[dat$area]

mu <- beta[1] + beta[2] * dat$x + dat$w_true
epsilon <- rnorm(nrow(dat), sd = sqrt(tau_sq))
dat$y_full <- mu + epsilon

holdout_id <- sample(seq_len(nrow(dat)), 6)
dat$sample <- "training"
dat$sample[holdout_id] <- "holdout"

dat$y <- dat$y_full
dat$y[holdout_id] <- NA
```

The CAR county-level smoothing and the east-west covariate trend are apparent in the mapped response values below.

```{r wa-response-map, fig.width=6.2, fig.height=3.8}
#| code-fold: true
#| code-summary: "Show plotting code"
wa_plot <- wa_counties
wa_plot$area <- as.character(wa_plot$COUNTYFIPS)
wa_plot$y_full <- dat$y_full[match(wa_plot$area, dat$area)]
wa_plot$sample <- dat$sample[match(wa_plot$area, dat$area)]

ggplot(wa_plot) +
  geom_sf(
    aes(fill = y_full),
    color = "white",
    linewidth = 0.15
  ) +
  geom_sf(
    data = wa_plot[wa_plot$sample == "holdout", ],
    fill = NA,
    color = "black",
    linewidth = 0.8
  ) +
  scale_fill_gradientn(colors = stlmm_palette()) +
  labs(fill = "response")
```

## Fit

Rows with `y = NA` are included in the fitting data so that the fitted graph still contains the held-out counties. They do not contribute to the likelihood, but they remain available for recovery and prediction because their county IDs are part of the CAR support.

The residual and process variances use half-$t$ priors on their corresponding standard deviations. The CAR association parameter `rho` uses a bounded uniform prior over positive spatial association values.

```{r fit-car}
fit <- stLMM(
  y ~ x + car(area, graph = g),
  data = dat,
  priors = list(
    resid = list(tau_sq = half_t(df = 3, scale = 0.5)),
    car_1 = list(
      sigma_sq = half_t(df = 3, scale = 1),
      rho = uniform(0.05, 0.95)
    )
  ),
  n_samples = 500,
  verbose = FALSE
)

summary(fit)
```

## Recover

Like other structured process terms, CAR effects are integrated out during fitting. Calling `recover()` draws the county-level CAR effects conditional on the posterior parameter samples.

```{r recover-car}
rec <- recover(fit, sub_sample = list(start = 150, thin = 2))

rec
```

The recovered `car_1` columns are ordered according to `g$ids`. As in the point-referenced spatial vignette, the data identify the combined areal intercept, $\beta_0 + w_i$, more directly than the split between the global intercept and the average level of one realized areal process. For this diagnostic, we therefore compare $\beta_0 + w_i$ rather than $w_i$ alone.

```{r recovery-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
rec_draws <- as_samples(rec, include_w = TRUE, metadata = FALSE)
w_cols <- paste0("w_car_1_", seq_along(g$ids))
w_hat <- colMeans(rec_draws[, w_cols, drop = FALSE])
names(w_hat) <- g$ids
beta_0_hat <- mean(rec_draws[["(Intercept)"]])

recovery_dat <- data.frame(
  truth = beta[1] + dat$w_true,
  estimate = beta_0_hat + w_hat[dat$area],
  sample = dat$sample
)
recovery_lim <- range(recovery_dat$truth, recovery_dat$estimate)

ggplot(recovery_dat, aes(truth, estimate)) +
  geom_point(
    aes(shape = sample),
    color = stlmm_color("primary"),
    size = 2.2
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(
    xlim = recovery_lim,
    ylim = recovery_lim
  ) +
  labs(
    x = "true areal intercept",
    y = "posterior mean estimate",
    shape = NULL
  )
```

## Predict

CAR prediction uses areas that already belong to the fitted graph. Here the held-out counties were included in the original data with `y = NA`, so they can be predicted after recovery.

```{r predict-holdout}
holdout_new <- dat[holdout_id, c("area", "x")]

pred <- predict(
  rec,
  newdata = holdout_new,
  y_samples = TRUE
)

summary(pred)
```

The `predict()` call returns posterior draws of $\mu$ in `mu_samples` and, because `y_samples = TRUE`, posterior predictive response draws in `y_samples`. Passing the prediction object to `summary()` gives posterior summaries for these draws.

```{r prediction-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
pred_y <- as_samples(pred, sample = "y", metadata = FALSE)
pred_dat <- data.frame(
  y = dat$y_full[holdout_id],
  y_hat = colMeans(pred_y)
)
pred_lim <- range(pred_dat$y, pred_dat$y_hat)

ggplot(pred_dat, aes(y, y_hat)) +
  geom_point(
    color = stlmm_color("primary"),
    size = 2.2
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(
    xlim = pred_lim,
    ylim = pred_lim
  ) +
  labs(
    x = "held-out response",
    y = "posterior predictive mean"
  )
```

## What this example illustrates

The `car()` term is for areal data where the latent process is indexed by graph nodes. The graph, not Euclidean distance, defines neighboring areas. Prediction is therefore different from point-referenced NNGP prediction: a CAR model can predict existing areas in the fitted graph, including areas whose responses were held out with `NA`, but it does not create new areas outside the graph.

## Choosing a CAR precision

The default `car()` term uses the package's original degree-scaled proper CAR precision. The same graph can also be used with the Leroux CAR precision [@LerouxLeiBreslow1999]:

```{r car-model-syntax, eval=FALSE}
y ~ x + car(area, graph = g, car_model = "proper")
y ~ x + car(area, graph = g, car_model = "leroux")
```

The Leroux option is useful when `rho` should describe a transition from nearly iid area effects to strong spatial smoothing. Both choices use the same `car_graph()` object, the same process variance parameter `sigma_sq`, and the same `rho` prior, starting, and tuning structure. The exact intrinsic-CAR endpoint `rho = 1` is not included; use an upper prior bound below one, such as `uniform(0.01, 0.99)`.
