---
title: "Spatially varying coefficients with stLMM"
subtitle: "Covariate-scaled NNGP terms, recovery, and prediction"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    code-overflow: wrap
    html-math-method: katex
execute:
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| code-fold: true
#| code-summary: "Show preliminaries"
library(stLMM)
library(ggplot2)

source(file.path(dirname(knitr::current_input(dir = TRUE)), "utils.R"))

set.seed(1)
```

```{r plot-theme, include=FALSE}
theme_set(theme_bw(base_size = 12))

show_table <- function(x, ...){
  knitr::kable(
    x,
    format = "html",
    escape = FALSE,
    table.attr = "class=\"stlmm-table\"",
    ...
  )
}
```

## Overview

Any structured process term in `stLMM` can be used as a covariate-scaled random effect. The formula syntax follows the usual interaction form. For example, `x:nngp(lon, lat)` gives a spatially varying coefficient for `x`, `x:ar1(day)` gives a time-varying coefficient, and `x:car(area, graph = g)` gives an areal spatially varying coefficient.

This vignette demonstrates the idea with an NNGP spatially varying coefficient. We simulate repeated observations at spatial locations, fit an NNGP model with both a spatially varying intercept and a spatially varying slope, recover the fitted coefficient surfaces, and predict held-out observations. The repeated observations are useful here because they give the model information to separate the local intercept from the local slope.

## Model

For observation $j$ at location $\mathbf{s}_i = (s_{i1}, s_{i2})^\top$, the model is

$$
y_{ij} = \beta_0 + w_0(\mathbf{s}_i)
       + x_{ij}\{\beta_1 + w_1(\mathbf{s}_i)\}
       + \epsilon_{ij},\qquad
\epsilon_{ij} \sim N(0,\tau^2).
$$

The two spatial processes are independent:

$$
\mathbf{w}_0 \sim N(\mathbf{0}, \sigma_0^2 \mathbf{R}_0(\phi_0)),
\qquad
\mathbf{w}_1 \sim N(\mathbf{0}, \sigma_1^2 \mathbf{R}_1(\phi_1)),
$$

with exponential correlation functions

$$
R_{k,ij}(\phi_k) = \exp\{-\phi_k \|\mathbf{s}_i - \mathbf{s}_j\|\},
\qquad k = 0, 1.
$$

The first process, $w_0(\mathbf{s})$, makes the intercept vary over space. The second process, $w_1(\mathbf{s})$, makes the coefficient for $x$ vary over space. The coefficient surface for $x$ is therefore $\beta_1 + w_1(\mathbf{s})$, not $w_1(\mathbf{s})$ alone.

## Data

We simulate locations in the unit square. At each location, we generate three observations with different values of `x`. The full Gaussian process covariance is used to simulate the two spatial surfaces. The fitted model uses `nngp()` as a nearest-neighbor approximation to the full process, as in the spatial NNGP vignette.

```{r simulate-data}
n_train_node <- 80
n_holdout_node <- 12
n_rep <- 3
n_node <- n_train_node + n_holdout_node

coords <- data.frame(
  lon = runif(n_node),
  lat = runif(n_node)
)

node <- rep(seq_len(n_node), each = n_rep)
dat <- data.frame(
  node = node,
  lon = coords$lon[node],
  lat = coords$lat[node]
)

dat$x <- rep(c(-1, 0, 1), times = n_node) + rnorm(n_node * n_rep, sd = 0.05)

beta <- c(0.7, 1.2)
tau_sq <- 0.05

sigma_sq_intercept <- 0.45
phi_intercept <- 5
sigma_sq_slope <- 0.45
phi_slope <- 4

C_intercept <- exp_cov(
  coords = coords,
  sigma_sq = sigma_sq_intercept,
  phi = phi_intercept
)
C_slope <- exp_cov(
  coords = coords,
  sigma_sq = sigma_sq_slope,
  phi = phi_slope
)

w_intercept <- rmvnorm(mean = rep(0, n_node), Sigma = C_intercept)
w_slope <- rmvnorm(mean = rep(0, n_node), Sigma = C_slope)

dat$intercept_true <- beta[1] + w_intercept[node]
dat$slope_true <- beta[2] + w_slope[node]
dat$mu <- dat$intercept_true + dat$x * dat$slope_true
dat$y <- dat$mu + rnorm(nrow(dat), sd = sqrt(tau_sq))

train <- dat[dat$node <= n_train_node, ]
holdout <- dat[dat$node > n_train_node, ]
```

```{r simulated-data-plot, fig.width=7, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
plot_nodes <- data.frame(
  lon = coords$lon[seq_len(n_train_node)],
  lat = coords$lat[seq_len(n_train_node)],
  intercept = beta[1] + w_intercept[seq_len(n_train_node)],
  slope = beta[2] + w_slope[seq_len(n_train_node)]
)

plot_dat <- rbind(
  data.frame(
    lon = plot_nodes$lon,
    lat = plot_nodes$lat,
    surface = "intercept",
    value = plot_nodes$intercept
  ),
  data.frame(
    lon = plot_nodes$lon,
    lat = plot_nodes$lat,
    surface = "slope for x",
    value = plot_nodes$slope
  )
)

ggplot(plot_dat, aes(lon, lat)) +
  geom_point(
    aes(color = value),
    size = 2.2
  ) +
  facet_wrap(~ surface) +
  scale_color_gradientn(colors = stlmm_palette()) +
  coord_equal() +
  labs(
    x = "longitude",
    y = "latitude",
    color = "true value"
  )
```

## Fit

The model formula includes `x` as a fixed effect and then adds two structured process terms. The first `nngp(lon, lat, ...)` term is the spatially varying intercept. The second term, `x:nngp(lon, lat, ...)`, is the spatially varying coefficient for `x`.

The two NNGP terms use separate covariance parameters. In the fitted object, the intercept process is named `nngp_1` and the spatially varying coefficient process is named `nngp_2`.

```{r fit-svc-nngp}
fit <- stLMM(
  y ~ x +
    nngp(lon, lat, m = 15, cov_model = "exp", ordering = "maxmin") +
    x:nngp(lon, lat, m = 15, cov_model = "exp", ordering = "maxmin"),
  data = train,
  priors = list(
    resid = list(tau_sq = half_t(df = 3, scale = 0.3)),
    nngp_1 = list(
      sigma_sq = half_t(df = 3, scale = 0.8),
      phi = uniform(1, 20)
    ),
    nngp_2 = list(
      sigma_sq = half_t(df = 3, scale = 0.8),
      phi = uniform(1, 20)
    )
  ),
  n_samples = 1500,
  verbose = FALSE
)

summary(fit)
```

## Recover

Structured process terms are integrated out during fitting. Calling `recover()` draws the fitted spatial process values conditional on the retained posterior samples.

```{r recover-svc}
rec <- recover(fit, sub_sample = list(start = 501, thin = 2))

rec
```

For direct `coda` workflows, `as_mcmc(rec, include_w = TRUE)` returns the
fixed-effect, covariance-parameter, and recovered coefficient-process samples
aligned to the recovered posterior iterations. The plotting code below uses
`as_samples()` because data-frame columns are easier to manipulate when
separating intercept and slope surfaces.

The recovered `nngp_1` values are the spatial intercept deviations $w_0(\mathbf{s}_i)$. The recovered `nngp_2` values are the spatial slope deviations $w_1(\mathbf{s}_i)$. To compare with the simulated coefficient surfaces, add the corresponding fixed-effect posterior mean.

Because the data have repeated observations at each training location, the recovered NNGP columns correspond to the unique training locations. We simulated and sorted the training data by node, so those columns align with `coords[1:n_train_node, ]`.

```{r recover-surfaces}
#| echo: false
rec_draws <- as_samples(rec, include_w = TRUE, metadata = FALSE)
w_intercept_cols <- paste0("w_nngp_1_", seq_len(n_train_node))
w_slope_cols <- paste0("w_nngp_2_", seq_len(n_train_node))
beta_hat <- colMeans(rec_draws[, c("(Intercept)", "x"), drop = FALSE])
w_intercept_hat <- colMeans(rec_draws[, w_intercept_cols, drop = FALSE])
w_slope_hat <- colMeans(rec_draws[, w_slope_cols, drop = FALSE])

recovery_dat <- data.frame(
  true_intercept = beta[1] + w_intercept[seq_len(n_train_node)],
  estimated_intercept = beta_hat["(Intercept)"] + w_intercept_hat,
  true_slope = beta[2] + w_slope[seq_len(n_train_node)],
  estimated_slope = beta_hat["x"] + w_slope_hat
)
```

```{r intercept-recovery-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
intercept_lim <- range(recovery_dat$true_intercept, recovery_dat$estimated_intercept)

ggplot(recovery_dat, aes(true_intercept, estimated_intercept)) +
  geom_point(
    color = stlmm_color("primary"),
    size = 2
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(
    xlim = intercept_lim,
    ylim = intercept_lim
  ) +
  labs(
    x = "true spatially varying intercept",
    y = "posterior mean estimate"
  )
```

```{r slope-recovery-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
slope_lim <- range(recovery_dat$true_slope, recovery_dat$estimated_slope)

ggplot(recovery_dat, aes(true_slope, estimated_slope)) +
  geom_point(
    color = stlmm_color("accent"),
    size = 2
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(
    xlim = slope_lim,
    ylim = slope_lim
  ) +
  labs(
    x = "true spatially varying slope",
    y = "posterior mean estimate"
  )
```

## Predict

Prediction works the same way as in the spatial NNGP vignette, except the new data must include the covariate used in the spatially varying coefficient term. The predicted mean combines the fixed effects, the predicted intercept process, and the predicted slope process multiplied by the new `x` value.

```{r predict-holdout}
holdout_new <- holdout[, c("lon", "lat", "x")]

pred <- predict(
  rec,
  newdata = holdout_new,
  y_samples = TRUE,
  joint = FALSE,
  return_w_samples = TRUE
)

pred
```

When `return_w_samples = TRUE`, the retained process samples include the raw predicted slope deviations $w_1(\mathbf{s})$. They do not include the covariate scaling. The scaling by `x` is applied when constructing the predicted means.

```{r prediction-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
pred_draws <- as_samples(pred, sample = "all", metadata = FALSE)
mu_cols <- paste0("mu_", seq_len(nrow(holdout_new)))
y_cols <- paste0("y_", seq_len(nrow(holdout_new)))
pred_dat <- data.frame(
  mu = holdout$mu,
  mu_hat = colMeans(pred_draws[, mu_cols, drop = FALSE]),
  y = holdout$y,
  y_hat = colMeans(pred_draws[, y_cols, drop = FALSE])
)

pred_lim <- range(pred_dat$mu, pred_dat$mu_hat)

ggplot(pred_dat, aes(mu, mu_hat)) +
  geom_point(
    color = stlmm_color("primary"),
    size = 2
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(
    xlim = pred_lim,
    ylim = pred_lim
  ) +
  labs(
    x = "true held-out mean",
    y = "posterior mean prediction"
  )
```

## What this example illustrates

The formula `x:nngp(lon, lat)` does not add another fixed effect for `x`; it adds a spatial process whose contribution is multiplied by `x`. Including both `x` and `x:nngp(lon, lat)` gives a global average slope plus a spatial deviation from that slope.

The same pattern applies to other structured terms. A covariate can scale an `ar1()` term for a time-varying coefficient, a `car()` term for an areal spatially varying coefficient, or a `car_time()` term for a space-time varying coefficient. The interpretation is the same: recover the raw latent process, then add the corresponding fixed effect when you want the full coefficient surface.

## Graphs and Sparse Factorization

One useful design detail is visible if we ask `stLMM` to print its term description before sampling. Setting `describe_terms = TRUE` prints the fixed-effect structure, process terms, graph indices, sparse matrix size, and CHOLMOD factorization diagnostics. The same option can be used with the model above; the example below uses fewer retained samples only to keep the vignette quick.

```{r describe-terms}
fit_described <- stLMM(
  y ~ x +
    nngp(lon, lat, m = 15, cov_model = "exp", ordering = "maxmin") +
    x:nngp(lon, lat, m = 15, cov_model = "exp", ordering = "maxmin"),
  data = train,
  priors = list(
    resid = list(tau_sq = half_t(df = 3, scale = 0.3)),
    nngp_1 = list(
      sigma_sq = half_t(df = 3, scale = 0.8),
      phi = uniform(1, 20)
    ),
    nngp_2 = list(
      sigma_sq = half_t(df = 3, scale = 0.8),
      phi = uniform(1, 20)
    )
  ),
  n_samples = 5,
  describe_terms = TRUE,
  verbose = FALSE
)
```

The output shows two process terms but only one graph. The first process term, `nngp_1`, is the spatially varying intercept and is not an SVC term. The second process term, `nngp_2`, is the `x:nngp(...)` term and is marked as an SVC term. Both terms have `graph index: 1`, which means they reuse the same NNGP graph: the same spatial support, neighbor count, ordering, and neighbor pattern. They still have separate covariance parameters and separate latent process values. Reusing the graph avoids rebuilding an identical nearest-neighbor structure for each term.

The model summary also reports the sparse matrix used by the collapsed sampler. Here `M dimensions` is the dimension of the latent sparse matrix $\mathbf{M}$, and `M nnz` is the number of stored nonzero entries in its sparse pattern. NNGP terms are useful computationally because each node is connected only to a small neighbor set, so the prior precision contribution is sparse.

After the sparse pattern is built, `stLMM` uses CHOLMOD, a sparse Cholesky factorization library for sparse symmetric positive definite matrices [@ChenDavisHagerRajamanickam2008CHOLMOD], through the R `Matrix` package's sparse matrix interface [@BatesMaechlerJagan2025Matrix]. CHOLMOD performs a symbolic analysis of the sparsity pattern and chooses a fill-reducing ordering; in this run the reported ordering is `amd`, approximate minimum degree. This ordering is internal to the factorization. It does not change the spatial coordinates, the NNGP neighbor graph, or the statistical model. Its job is to reduce fill-in, meaning new nonzero entries created in the Cholesky factor that were not present in the original sparse matrix pattern. If CHOLMOD is compiled with METIS support, it can also use graph-partitioning orderings based on multilevel partitioning ideas [@KarypisKumar1998SparseOrdering], which can reduce fill-in and factorization work for some sparse matrix patterns. Once the symbolic structure is available, later numerical factorizations can reuse that structure as covariance and variance parameters change.

The remaining CHOLMOD diagnostics are computational, not statistical. `CHOLMOD fill ratio` compares the size of the factor pattern to the original sparse matrix pattern. `CHOLMOD lnz` reports the number of nonzero entries in the Cholesky factor, and `CHOLMOD flops` reports the approximate floating-point work for the factorization. These quantities help explain why graph structure, repeated-location collapsing, shared graph reuse, and the NNGP neighbor count `m` can matter for runtime.

This sparse-matrix view is central to the package design. NNGP, CAR, and autoregressive terms all produce sparse precision matrices, and `stLMM` is built around using that sparsity with CHOLMOD sparse factorization rather than treating the latent process covariance as dense. The lower-level dense linear algebra used inside these routines is handled by BLAS and LAPACK [@BlackfordEtAl2002BLAS; @AndersonEtAl1999LAPACK]. Because those libraries are supplied by the user's R installation, the same package code can benefit from optimized or threaded BLAS/LAPACK implementations when R is linked against them. One common open-source choice is OpenBLAS [@OpenBLAS2025], though the best option depends on the user's operating system, processor, and R installation. For a deeper look at CHOLMOD ordering, fill-in diagnostics, OpenMP threads, BLAS/LAPACK choices, and reproducible timing scripts, see the [runtime-performance article](a11-runtime-performance.html).
