---
title: "Negative-binomial count models with stLMM"
subtitle: "Fixed-size NB-PG likelihoods with areal CAR effects"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    code-overflow: wrap
    html-math-method: katex
execute:
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| code-fold: true
#| code-summary: "Show preliminaries"
library(stLMM)
library(ggplot2)
library(sf)

source(file.path(dirname(knitr::current_input(dir = TRUE)), "utils.R"))

set.seed(14)
```

```{r plot-theme, include=FALSE}
theme_set(theme_bw(base_size = 12))
chain_colors <- stlmm_discrete_colors(3)
trace_parameters <- c("(Intercept)", "x", "car_1_sigma_sq", "car_1_rho")
```

## Overview

This vignette introduces count models in `stLMM` using `family = "negative_binomial"`. The example uses the bundled `stunitco` county polygon layer, extracts Washington counties, builds a county adjacency graph, simulates areal counts with a CAR latent effect and a known exposure offset, and fits fixed-size negative-binomial models using Pólya-Gamma updates. The fits use three chains so the trace plots can be inspected directly.

There are two simulation sections. The first generates counts from a negative-binomial model with known size and fits the same likelihood. The second generates Poisson counts and fits a large-size negative-binomial model to show how the fixed-size NB likelihood can approximate a Poisson log-mean model. The examples avoid extremely sparse counts, because very sparse count data provide little information about a latent areal surface.

## Count model

For area $i$, `stLMM` uses the negative-binomial mean parameterization

$$
y_i \mid \mu_i, r \sim \operatorname{NB}(r, \mu_i),
\qquad
E(y_i) = \mu_i,
\qquad
\operatorname{var}(y_i) = \mu_i + \frac{\mu_i^2}{r},
$$

where $r$ is the fixed `size` argument. The log mean is modeled as

$$
\eta_i = \log(\mu_i) = o_i + \beta_0 + x_i\beta_1 + w_i.
$$

Here $o_i$ is a known formula offset. In the simulations below, $o_i = \log(e_i)$ for a known exposure multiplier $e_i$.

Internally, the Pólya-Gamma augmentation uses the logit probability tilt

$$
\psi_i = \eta_i - \log(r),
$$

and draws

$$
\omega_i \mid \psi_i, y_i, r \sim \operatorname{PG}(y_i + r, \psi_i).
$$

Conditional on $\omega_i$, the likelihood is Gaussian in $\eta_i$ with working response

$$
z_i = \frac{(y_i-r)/2}{\omega_i} + \log(r).
$$

This is the same Pólya-Gamma identity used for binomial logistic regression, applied to the negative-binomial logit-probability representation [@PolsonScottWindle2013; @WindleCarvalhoScottSun2013]. The package uses this conditionally Gaussian form to reuse the sparse collapsed sampler used by the Gaussian and binomial models.

The `size` parameter is fixed in this implementation. A larger `size` reduces overdispersion and moves the negative-binomial distribution toward a Poisson distribution. Extremely large `size` values are not always better computationally because the Pólya-Gamma shape is $y_i + r$.

## County graph

We use the same Washington county graph as the areal CAR vignette. The graph defines the latent areal support and the neighbor relationships used by `car()`.

```{r wa-data}
data(stunitco, package = "stLMM")

wa_counties <- stunitco[stunitco$STATECD == 53, ]
g <- car_graph(wa_counties, id = "COUNTYFIPS", island = "nearest", island_k = 4)
```

```{r wa-edge-graph, fig.width=6.2, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show edge-map code"
coord_dat <- data.frame(
  area = as.character(wa_counties$COUNTYFIPS),
  sf::st_coordinates(
    sf::st_point_on_surface(sf::st_geometry(wa_counties))
  )
)

edge_index <- which(as.matrix(g$adjacency) != 0, arr.ind = TRUE)
edge_index <- edge_index[edge_index[, "row"] < edge_index[, "col"], , drop = FALSE]
edge_dat <- data.frame(
  from = g$ids[edge_index[, "row"]],
  to = g$ids[edge_index[, "col"]]
)
edge_dat$x <- coord_dat$X[match(edge_dat$from, coord_dat$area)]
edge_dat$y <- coord_dat$Y[match(edge_dat$from, coord_dat$area)]
edge_dat$xend <- coord_dat$X[match(edge_dat$to, coord_dat$area)]
edge_dat$yend <- coord_dat$Y[match(edge_dat$to, coord_dat$area)]

ggplot(wa_counties) +
  geom_sf(fill = "grey95", color = "white", linewidth = 0.15) +
  geom_segment(
    data = edge_dat,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "grey45",
    linewidth = 0.25
  ) +
  geom_point(
    data = coord_dat,
    aes(X, Y),
    color = stlmm_color("primary"),
    size = 1.4
  ) +
  labs(x = "longitude", y = "latitude")
```

## Negative-binomial data

We create a county covariate from east-west position, define a known exposure multiplier, and simulate a smooth areal effect from the proper CAR precision. Counts are generated from the negative-binomial model with fixed size $r=12$. Four counties are held out by setting their observed response to `NA`; the rows stay in the data so their areal effects can be recovered and predicted.

```{r simulate-nb-data}
centroid_x <- sf::st_coordinates(sf::st_centroid(sf::st_geometry(wa_counties)))[, "X"]

dat_nb <- data.frame(
  area = as.character(wa_counties$COUNTYFIPS),
  county = wa_counties$COUNTYNM,
  x = as.numeric(scale(centroid_x))
)
dat_nb$log_exposure <- 0.25 * dat_nb$x

beta_nb <- c("(Intercept)" = 2.5, x = -0.35)
size_nb <- 12
sigma_sq_nb <- 1
rho_nb <- 0.9

Q_nb <- car_prec(g, sigma_sq = sigma_sq_nb, rho = rho_nb)
Sigma_nb <- solve(Q_nb)
w_nb <- rmvnorm(mean = rep(0, nrow(Q_nb)), Sigma = Sigma_nb)
names(w_nb) <- rownames(Q_nb)

dat_nb$w_true <- w_nb[dat_nb$area]
dat_nb$eta_true <- dat_nb$log_exposure +
  beta_nb["(Intercept)"] + beta_nb["x"] * dat_nb$x + dat_nb$w_true
dat_nb$mu_true <- exp(dat_nb$eta_true)
dat_nb$y_full <- rnbinom(nrow(dat_nb), size = size_nb, mu = dat_nb$mu_true)

holdout_nb <- sample(seq_len(nrow(dat_nb)), 4)
dat_nb$sample <- "training"
dat_nb$sample[holdout_nb] <- "holdout"
dat_nb$y <- dat_nb$y_full
dat_nb$y[holdout_nb] <- NA
```

```{r nb-response-map, fig.width=6.2, fig.height=3.8}
#| code-fold: true
#| code-summary: "Show plotting code"
wa_nb <- wa_counties
wa_nb$area <- as.character(wa_nb$COUNTYFIPS)
wa_nb$y_full <- dat_nb$y_full[match(wa_nb$area, dat_nb$area)]
wa_nb$sample <- dat_nb$sample[match(wa_nb$area, dat_nb$area)]

ggplot(wa_nb) +
  geom_sf(aes(fill = y_full), color = "white", linewidth = 0.15) +
  geom_sf(
    data = wa_nb[wa_nb$sample == "holdout", ],
    fill = NA,
    color = "black",
    linewidth = 0.8
  ) +
  scale_fill_gradientn(colors = stlmm_palette()) +
  labs(fill = "count")
```

## Fit the NB model

The `family = "negative_binomial"` argument selects the count likelihood, and `size = size_nb` fixes the negative-binomial size. There is no Gaussian residual variance term. The observation-level variability is defined by the count likelihood and the fixed size.

```{r fit-nb}
fit_nb <- stLMM(
  y ~ x + offset(log_exposure) + car(area, graph = g),
  data = dat_nb,
  family = "negative_binomial",
  size = size_nb,
  priors = list(
    beta = normal(mean = 0, sd = 3),
    car_1 = list(
      sigma_sq = half_t(df = 3, scale = 1),
      rho = uniform(0.05, 0.95)
    )
  ),
  starting = list(car_1 = list(sigma_sq = 1, rho = 0.8)),
  tuning = list(car_1 = list(sigma_sq = 0.08, rho = 0.06)),
  n_samples = 1500,
  chains = 3,
  chain_control = list(seed = 14, dispersion = 1.5),
  verbose = FALSE
)

summary(fit_nb)
```

```{r nb-chain-traces, fig.width=8, fig.height=5}
#| code-fold: true
#| code-summary: "Show trace-plot code"
plot(
  fit_nb,
  parameters = trace_parameters,
  n_col = 2,
  burnin = 500,
  chain_colors = chain_colors
)
```

## Recover areal effects

As with Gaussian CAR models, the structured CAR effect is integrated out during parameter updates. For Pólya-Gamma process models, `stLMM()` saves the in-chain process draws by default as `save_process = list(start = 1, thin = 1)` because those draws are already needed to refresh the augmentation variables. Calling `recover()` selects the retained county-effect draws requested by `sub_sample`; it does not run a separate post-fit Pólya-Gamma reconstruction. Use `save_process = list(start = ..., thin = ...)` to reduce memory use for larger process supports.

The intercept and the average level of a realized spatial effect can trade off, so the diagnostic below compares the areal intercept $\beta_0 + w_i$ rather than $w_i$ alone. The full log mean $\eta_i = \beta_0 + x_i\beta_1 + w_i$ is often the more stable prediction-scale diagnostic, but $\beta_0 + w_i$ is useful for seeing whether the structured surface is being recovered.

```{r recover-nb}
rec_nb <- recover(fit_nb, sub_sample = list(start = 501, thin = 5))
rec_nb_draws <- as_samples(rec_nb, include_w = TRUE, metadata = FALSE)

w_nb_cols <- paste0("w_car_1_", seq_along(g$ids))
w_nb_hat <- colMeans(rec_nb_draws[, w_nb_cols, drop = FALSE])
names(w_nb_hat) <- g$ids
beta_nb_hat <- mean(rec_nb_draws[["(Intercept)"]])

nb_recovery <- data.frame(
  truth = beta_nb["(Intercept)"] + dat_nb$w_true,
  estimate = beta_nb_hat + w_nb_hat[dat_nb$area],
  sample = dat_nb$sample
)
head(nb_recovery)
```

```{r nb-recovery-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
nb_recovery_lim <- range(nb_recovery$truth, nb_recovery$estimate)

ggplot(nb_recovery, aes(truth, estimate)) +
  geom_point(
    aes(shape = sample),
    color = stlmm_color("primary"),
    size = 2.2
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(xlim = nb_recovery_lim, ylim = nb_recovery_lim) +
  labs(
    x = "true areal intercept",
    y = "posterior mean estimate",
    shape = NULL
  )
```

## Fitted means and holdout counts

The default response scale for negative-binomial models is the mean count $\mu_i = \exp(\eta_i)$. The link scale is the log mean.

```{r fitted-nb}
fitted_nb <- fitted(rec_nb)
fitted_nb_link <- fitted(rec_nb, scale = "link")

head(data.frame(mean_count = fitted_nb, log_mean = fitted_nb_link))
```

```{r predict-nb}
holdout_nb_new <- dat_nb[holdout_nb, c("area", "x", "log_exposure")]

pred_nb <- predict(
  rec_nb,
  newdata = holdout_nb_new,
  y_samples = TRUE
)

summary(pred_nb)
```

```{r nb-prediction-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
pred_nb_mu <- as_samples(pred_nb, sample = "mu", metadata = FALSE)
pred_nb_dat <- data.frame(
  mu = dat_nb$mu_true[holdout_nb],
  mu_hat = colMeans(pred_nb_mu),
  y = dat_nb$y_full[holdout_nb]
)
pred_nb_lim <- range(pred_nb_dat$mu, pred_nb_dat$mu_hat)

ggplot(pred_nb_dat, aes(mu, mu_hat)) +
  geom_point(aes(size = y), color = stlmm_color("primary"), alpha = 0.85) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(xlim = pred_nb_lim, ylim = pred_nb_lim) +
  labs(
    x = "true held-out mean",
    y = "posterior mean prediction",
    size = "held-out count"
  )
```

## Poisson data and a large-size approximation

If $r$ is large, the negative-binomial variance $\mu_i + \mu_i^2/r$ is close to the Poisson variance $\mu_i$. This section simulates Poisson counts on the same county support and fits a negative-binomial model with a large fixed size. This is useful as a diagnostic for the approximation, not a replacement for a dedicated Poisson likelihood.

```{r simulate-poisson-data}
set.seed(1)

dat_pois <- dat_nb[, c("area", "county", "x", "log_exposure")]

beta_pois <- c("(Intercept)" = 5, x = -0.5)
sigma_sq_pois <- 1
rho_pois <- 0.9
size_pois_approx <- 1000

Q_pois <- car_prec(g, sigma_sq = sigma_sq_pois, rho = rho_pois)
Sigma_pois <- solve(Q_pois)
w_pois <- rmvnorm(mean = rep(0, nrow(Q_pois)), Sigma = Sigma_pois)
names(w_pois) <- rownames(Q_pois)

dat_pois$w_true <- w_pois[dat_pois$area]
dat_pois$eta_true <- dat_pois$log_exposure +
  beta_pois["(Intercept)"] + beta_pois["x"] * dat_pois$x + dat_pois$w_true
dat_pois$mu_true <- exp(dat_pois$eta_true)
dat_pois$y <- rpois(nrow(dat_pois), lambda = dat_pois$mu_true)
```

```{r poisson-response-map, fig.width=6.2, fig.height=3.8}
#| code-fold: true
#| code-summary: "Show plotting code"
wa_pois <- wa_counties
wa_pois$area <- as.character(wa_pois$COUNTYFIPS)
wa_pois$y <- dat_pois$y[match(wa_pois$area, dat_pois$area)]

ggplot(wa_pois) +
  geom_sf(aes(fill = y), color = "white", linewidth = 0.15) +
  scale_fill_gradientn(colors = stlmm_palette()) +
  labs(fill = "count")
```

```{r fit-poisson-approx}
fit_pois_nb <- stLMM(
  y ~ x + offset(log_exposure) + car(area, graph = g),
  data = dat_pois,
  family = "negative_binomial",
  size = size_pois_approx,
  priors = list(
    beta = normal(mean = 0, sd = 3),
    car_1 = list(
      sigma_sq = half_t(df = 3, scale = 1),
      rho = uniform(0.05, 0.95)
    )
  ),
  starting = list(car_1 = list(sigma_sq = 1, rho = 0.8)),
  tuning = list(car_1 = list(sigma_sq = 0.08, rho = 0.06)),
  n_samples = 1200,
  chains = 3,
  chain_control = list(seed = 15, dispersion = 1.5),
  verbose = FALSE
)

summary(fit_pois_nb)
```

```{r poisson-approx-chain-traces, fig.width=8, fig.height=5}
#| code-fold: true
#| code-summary: "Show trace-plot code"
plot(
  fit_pois_nb,
  parameters = trace_parameters,
  n_col = 2,
  burnin = 400,
  chain_colors = chain_colors
)
```

```{r recover-poisson-approx}
rec_pois_nb <- recover(fit_pois_nb, sub_sample = list(start = 401, thin = 4))
rec_pois_draws <- as_samples(rec_pois_nb, include_w = TRUE, metadata = FALSE)

w_pois_hat <- colMeans(rec_pois_draws[, w_nb_cols, drop = FALSE])
names(w_pois_hat) <- g$ids
beta_pois_hat <- mean(rec_pois_draws[["(Intercept)"]])

pois_recovery <- data.frame(
  truth = beta_pois["(Intercept)"] + dat_pois$w_true,
  estimate = beta_pois_hat + w_pois_hat[dat_pois$area]
)
```

```{r poisson-approx-plot, fig.width=5.5, fig.height=4.8}
#| code-fold: true
#| code-summary: "Show plotting code"
pois_recovery_lim <- range(pois_recovery$truth, pois_recovery$estimate)

ggplot(pois_recovery, aes(truth, estimate)) +
  geom_point(color = stlmm_color("primary"), size = 2.2) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = stlmm_color("secondary"),
    linewidth = 0.8
  ) +
  coord_equal(xlim = pois_recovery_lim, ylim = pois_recovery_lim) +
  labs(
    x = "true areal intercept",
    y = "posterior mean estimate"
  )
```

The large-size fit uses the same package machinery as the negative-binomial example. In this small simulation it should recover the log mean surface reasonably well, but the approximation is still an NB model. In applications where a Poisson likelihood is the scientific target, a native Poisson implementation would be the cleaner model.

## Practical guidance

The fixed-size negative-binomial family is useful when counts are overdispersed relative to Poisson and a fixed size is scientifically defensible or can be sensitivity-tested. A simple workflow is to fit a small grid of plausible sizes and compare fixed effects, fitted means, spatial variance, and recovered areal intercepts.

The `size` parameter also interacts with random and structured effects. Smaller `size` gives extra observation-level overdispersion, while spatial random effects explain persistent area-level variation. Because these components can trade off, report sensitivity to `size` when the data do not strongly determine the distinction.

Very large `size` values approximate Poisson counts, but they are not automatically better. The Pólya-Gamma draw has shape $y_i + r$, so large `size` can increase computational cost or change sampler behavior. Use a large-size NB fit as an approximation check, not as a blanket recommendation.

For negative-binomial models, `fitted()` and `predict()` return mean counts on the response scale by default. Use `scale = "link"` for the log mean. If `y_samples = TRUE`, prediction simulates negative-binomial count outcomes using the fixed `size`.
