---
title: "Metropolis blocking and adaptation in stLMM"
subtitle: "How covariance and residual-variance proposals are tuned, blocked, and diagnosed"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    code-overflow: wrap
    html-math-method: katex
execute:
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| code-fold: true
#| code-summary: "Show preliminaries"
library(stLMM)
library(ggplot2)

source(file.path(dirname(knitr::current_input(dir = TRUE)), "utils.R"))

set.seed(1)
```

```{r plot-theme, include=FALSE}
theme_set(theme_bw(base_size = 12))

show_table <- function(x, ...){
  knitr::kable(
    x,
    format = "html",
    escape = FALSE,
    table.attr = "class=\"stlmm-table\"",
    ...
  )
}
```

## Overview

This vignette focuses on the Metropolis part of the `stLMM` sampler: which parameters enter it, how proposals are adapted, what the blocking options do, and how to decide whether a blocking scheme is worth the added computing time.

The discussion is about the sampler implemented in `stLMM`, not a claim that the collapsed parameterization is best for every model or even the most natural parameterization for blocking. Similar models, and much more sophisticated hierarchical models, can be fit in general Bayesian modeling software. Those tools can represent the model in different ways: latent variables may be retained as parameters, centered or non-centered parameterizations may be used, and groups of model nodes or parameters may be updated with customized samplers. Those choices can be more favorable for a particular posterior geometry [@PapaspiliopoulosRobertsSkold2007; @StanUserGuideReparameterization]. Software built around programmable model graphs, such as NIMBLE, exposes this kind of sampler customization and even automated blocking because useful blocks are model- and parameterization-specific [@deValpineEtAl2017NIMBLE; @NimbleManualMCMC].

The useful role for `stLMM` is narrower. It gives formula-level access to a specific family of spatial and space-time linear mixed models, including structured covariance terms, NNGP approximations, residual-variance options used in small area estimation, recovery and prediction methods, and fixed-parameter controls. The aim is to make these models easier to fit without requiring users to write the full latent model and sampler themselves. The collapsed sampler reduces the Markov chain dimension and gives one coherent implementation across these terms, but blocking then operates only on the covariance and residual-variance parameters through the collapsed target. That can be an advantage for routine models and a limitation for difficult posterior geometries.

`stLMM` uses a collapsed sampler for structured latent process terms such as `ar1()`, `gp()`, `nngp()`, `car()`, and `car_time()`. The uncollapsed Gaussian model can be written as

$$
  \boldsymbol{y}
    = \mathbf{X}\boldsymbol{\beta}
      + \mathbf{Z}\boldsymbol{\alpha}
      + \mathbf{A}\boldsymbol{w}
      + \boldsymbol{\varepsilon},
$$

$$
  \boldsymbol{w} \sim N(\boldsymbol{0}, \mathbf{Q}_w^{-1}),
  \qquad
  \boldsymbol{\varepsilon} \sim N(\boldsymbol{0}, \mathbf{R}),
$$

where $\mathbf{A}$ maps observation rows to structured latent process nodes, $\mathbf{Q}_w$ is the structured-process prior precision, $\mathbf{R}$ is the diagonal residual covariance, and $\mathbf{W}=\mathbf{R}^{-1}$ is the diagonal observation precision. Under the default homoskedastic Gaussian residual model, $\mathbf{R}=\tau^2\mathbf{I}_n$ and $\mathbf{W}=\tau^{-2}\mathbf{I}_n$.

During fitting, the structured process vector $\boldsymbol{w}$ isn't sampled while covariance parameters are updated. Instead, $\boldsymbol{w}$ is integrated out. We reserve $\boldsymbol{\theta}_k$ for the covariance-function parameters of structured process term $k$, such as a range, decay, correlation, smoothness, or nonseparable mixing parameter. The broader Metropolis-updated parameter collection is denoted $\boldsymbol{\psi}$; it can include $\tau^2$, residual variances, process variances $\sigma_k^2$, and $\boldsymbol{\theta}_k$ values. Conditional on $\boldsymbol{\beta}$, $\boldsymbol{\alpha}$, and $\boldsymbol{\psi}$, the collapsed observation model is

$$
  \boldsymbol{y} \mid
    \boldsymbol{\beta}, \boldsymbol{\alpha}, \boldsymbol{\psi}
  \sim
  N(\mathbf{X}\boldsymbol{\beta}
      + \mathbf{Z}\boldsymbol{\alpha}, \mathbf{V}),
$$

with

$$
  \mathbf{V}
    = \mathbf{R} + \mathbf{A}\mathbf{Q}_w^{-1}\mathbf{A}^{\top}.
$$

The code doesn't form $\mathbf{V}$ explicitly; rather, it evaluates the collapsed target through the sparse latent matrix

$$
  \mathbf{M} = \mathbf{Q}_w + \mathbf{A}^{\top}\mathbf{W}\mathbf{A},
$$

which gives

$$
\begin{aligned}
  \mathbf{V}^{-1}
    &= \mathbf{W}
       - \mathbf{W}\mathbf{A}\mathbf{M}^{-1}
         \mathbf{A}^{\top}\mathbf{W}, \\
  \log|\mathbf{V}|
    &= \log|\mathbf{R}| + \log|\mathbf{M}| - \log|\mathbf{Q}_w|.
\end{aligned}
$$

The dependence of $\mathbf{R}$, $\mathbf{W}$, $\mathbf{Q}_w$, $\mathbf{V}$, and $\mathbf{M}$ on the current value of $\boldsymbol{\psi}$ is suppressed to keep the notation readable.

This collapsed-sampler choice is deliberate, but it has real costs. It reduces the retained MCMC parameter space and lets `stLMM` fit several structured random-effect models in one framework, including dense GP, NNGP, AR(1), CAR, CAR-time, and covariate-scaled process terms. Once $\boldsymbol{w}$ is integrated out, however, many covariance and variance parameters don't have simple Gibbs updates. A proposed range, correlation, mixing, process-variance, or residual-variance value has to be evaluated through the collapsed target, which can require rebuilding covariance or precision pieces, assembling $\mathbf{M}$, and factorizing $\mathbf{M}$.

Blocking is the choice of which Metropolis-updated parameters are proposed together. One large block can be efficient when posterior dependence is well captured by the adaptive proposal covariance, but it can struggle when different parameter groups move on different scales or have different posterior dependence patterns. Splitting parameters into smaller blocks can sometimes improve mixing because each proposal is tuned to a more focused part of the posterior. There is no free lunch: each extra Metropolis block can add another target evaluation and factorization within the same retained iteration.

After collapsing the structured process terms, only a few parameter groups have Gibbs updates:

- fixed effects $\boldsymbol{\beta}$ are updated by Gibbs steps;
- unstructured grouped random effects $\boldsymbol{\alpha}$, when included through terms such as `iid(group)`, are updated by Gibbs steps;
- structured process effects $\boldsymbol{w}$ are recovered after fitting with `recover()`, not sampled during covariance-parameter updates.

The default Metropolis update is one joint block containing the covariance and residual-variance parameters that are updated by Metropolis. That's usually fastest. Splitting the parameters into several blocks can improve mixing when one large proposal is hard to tune, but it can make each iteration more expensive. Blocking is therefore a computational tradeoff, not a universally better sampler.

The retained adaptive Metropolis strategy used in `stLMM` adapts ideas from unpublished adaptive Metropolis notes by Benjamin Shaby and from the broader adaptive MCMC literature [@andrieu2008tutorial].

`stLMM` records sampler timing in the returned fit object. For a single-chain fit, `fit$timing$sampler` gives the usual `proc.time()` CPU and elapsed wall-clock differences around the compiled sampler call. For a multi-chain fit, `fit$timing$sampler_by_chain` gives those values by chain and `fit$timing$sampler_total` gives the total across chains. We'll use those values below to compare the computational cost of blocking schemes.

## What Enters the Metropolis Update

The Metropolis machinery only sees covariance and residual-variance parameters that are updated during MCMC. Fixed parameters declared with `fixed(value)` are omitted from every Metropolis block. They stay at their fixed values and don't need priors.

The sampler proposes on transformed scales. Positive parameters, such as $\tau^2$ and process variances, are updated on the log scale. Bounded covariance-function parameters are updated on a bounded-logit scale. For example, in an exponential NNGP model with residual variance, process variance, and spatial decay all updated by Metropolis, one possible transformed vector is

$$
  \boldsymbol{\eta} =
  \{\log(\tau^2), \log(\sigma^2), g(\phi)\}^{\top},
$$

where $\phi$ is the term's decay parameter and $g(\cdot)$ maps its bounded prior support to the real line. In general, $\boldsymbol{\eta}$ contains the transformed parameters that are updated by Metropolis. Priors and transformation Jacobians are included in the Metropolis target, so proposals are made on convenient scales while the posterior remains the model specified by the user.

## The Adaptive Metropolis Step

Let $\boldsymbol{\eta}_b$ be the current transformed parameter vector for block $b$. All parameters outside block $b$ are held fixed while that block is proposed. Let $\ell(\boldsymbol{\eta}_b)$ denote the collapsed log posterior target under the current block values, including the collapsed likelihood, priors, and transformation Jacobians.

For each retained MCMC iteration, the sampler loops over the Metropolis blocks. For each block $b$:

1. Propose a random walk on the transformed scale,

$$
  \boldsymbol{\eta}_b^\star
    = \boldsymbol{\eta}_b + \mathbf{L}_b \boldsymbol{z},
  \qquad
  \boldsymbol{z} \sim N(\boldsymbol{0}, \mathbf{I}).
$$

For multivariate blocks, $\mathbf{L}_b\mathbf{L}_b^{\top}=\lambda_b\boldsymbol{\Sigma}_b$, where $\boldsymbol{\Sigma}_b$ is the adaptive empirical covariance shape and $\lambda_b$ is a variance-like multiplier. For scalar blocks, the proposal variance is $\delta_b^2\Sigma_b$, where $\Sigma_b>0$ is a scalar shape and $\delta_b$ is a one-dimensional random-walk scale. The two symbols reflect the implementation: multivariate blocks adapt an empirical covariance shape and a multiplier on that covariance, while scalar blocks adapt a direct step size.

2. Transform $\boldsymbol{\eta}_b^\star$ back to the model scale.

3. Rebuild the affected covariance or precision pieces, assemble $\mathbf{M}$, and factorize it. If the proposed covariance or precision isn't valid, reject the proposal.

4. Otherwise accept with probability

$$
  \min\left\{
    1,
    \exp\left[
      \ell(\boldsymbol{\eta}_b^\star)
      - \ell(\boldsymbol{\eta}_b)
    \right]
  \right\}.
$$

5. At the end of each retained adaptation batch, compute the block acceptance rate $\hat{a}_b$. For multivariate blocks, update the proposal multiplier and empirical covariance by

$$
  \log(\lambda_b) \leftarrow \log(\lambda_b) +
    c_0 (m+1)^{-c_1}(\hat{a}_b-a^\star),
$$

$$
  \boldsymbol{\Sigma}_b \leftarrow \boldsymbol{\Sigma}_b +
    (m+1)^{-c_1}
      (\widehat{\mathrm{Cov}}_b-\boldsymbol{\Sigma}_b),
$$

where $m$ is the completed batch index, $a^\star$ is the target acceptance rate, and the implementation uses $c_0=1$ and $c_1=0.8$.

Scalar blocking uses one-dimensional scale adaptation for $\delta_b$ and defaults to target acceptance 0.44. Multivariate blocking defaults to target acceptance 0.234. The adaptation is diminishing, so later batches make smaller changes. Diminishing adaptation is one of the standard conditions used in adaptive MCMC theory; paired with suitable control of the proposal kernels, it helps preserve ergodicity while allowing the proposal to keep learning during the run [@RobertsRosenthal2007Coupling; @Roberts01012009].

## Warmup and Retained Adaptation

There are two tuning phases, and they serve different purposes.

First, the warmup phase adjusts proposal scales before retained samples are collected. Warmup draws aren't returned as posterior samples, but warmup diagnostics are stored in `fit$adaptive_metropolis$warmup`. In the next phase, retained adaptation continues during the posterior run. Its diagnostics are stored in `fit$adaptive_metropolis` and, for blocked updates, in `fit$adaptive_metropolis$blocks`. See `?stLMM` for the full set of `warmup` and `metropolis` controls and their defaults.

## Blocking Choices

Blocking changes the proposal partition. It doesn't change the model, likelihood, prior, or posterior target.

```{r blocking-table}
#| echo: false
blocking_table <- data.frame(
  Blocking = c(
    "\"joint\"",
    "\"by_term\"",
    "\"residual_process\"",
    "\"variance_theta\"",
    "\"process_variance\"",
    "\"scalar\""
  ),
  Partition = c(
    "All covariance and residual-variance parameters updated by Metropolis in one block.",
    "Residual variance in one block; each structured process term's variance and covariance-function parameters in its own block.",
    "Residual variances in one block; process covariance parameters in another.",
    "Variance parameters, including residual and process variances, in one block; bounded covariance parameters in another.",
    "Residual variance in one block; process variances in one block; bounded covariance parameters in one block.",
    "Each covariance or residual-variance parameter is updated individually in its own one-dimensional block."
  ),
  Use = c(
    "Start here; usually fastest.",
    "Several process terms appear to tune differently.",
    "Residual variances dominate the joint proposal.",
    "Variance-like parameters and bounded covariance-function parameters move differently.",
    "Residual variance, process variance, and covariance-function parameters each need separate movement.",
    "Diagnostic fallback when larger blocks don't move."
  ),
  check.names = FALSE
)

show_table(blocking_table)
```

### A Diagnostic NNGP Example

The example below uses one spatial NNGP term. The Metropolis-updated parameters are the residual variance `tau_sq`, the process variance `nngp_1_sigma_sq`, and the spatial decay `nngp_1_phi`.

This is a useful diagnostic example because the two variance parameters can move differently from the decay parameter. We'll compare:

- `"joint"`, the default, which updates all three parameters together;
- `"by_term"`, which updates `tau_sq` by itself, then updates `nngp_1_sigma_sq` and `nngp_1_phi` together.

```{r simulate-data}
n <- 200

dat <- data.frame(
  x_coord = runif(n),
  y_coord = runif(n),
  x = runif(n, -1, 1)
)

beta <- c(0.5, 1)
tau_sq <- 0.1
sigma_sq <- 0.5
phi <- 5

C <- exp_cov(
  dat[, c("x_coord", "y_coord")],
  sigma_sq = sigma_sq,
  phi = phi
)

w <- rmvnorm(mean = rep(0, n), Sigma = C)
mu <- beta[1] + beta[2] * dat$x + w
dat$y <- mu + rnorm(n, sd = sqrt(tau_sq))
```

The chains are short enough for a vignette, but long enough to show the blocking tradeoff without relying on a specially selected data realization. We set `chain_control$dispersion` above its default to start the chains farther apart for the positive variance parameters. In a real analysis, use longer chains and don't make a blocking decision from one run alone.

```{r fit-helper}
fit_with_blocking <- function(blocking){
  stLMM(
    y ~ x + nngp(x_coord, y_coord, m = 15,
                 cov_model = "exp", ordering = "maxmin"),
    data = dat,
    priors = list(
      resid = list(tau_sq = half_t(df = 3, scale = 0.5)),
      nngp_1 = list(
        sigma_sq = half_t(df = 3, scale = 1),
        phi = uniform(1, 15)
      )
    ),
    n_samples = 1500,
    chains = 3,
    chain_control = list(seed = 1, dispersion = 2),
    metropolis = list(
      blocking = blocking,
      batch_length = 25
    ),
    verbose = FALSE
  )
}
```

```{r fit-blocking}
joint_run <- fit_with_blocking("joint")
by_term_run <- fit_with_blocking("by_term")
```

### Compare Blocking Diagnostics

First check the blocks used by the first chain. The table reports the model-scale parameter names so the grouping is easy to read.

```{r block-summary}
#| echo: false
display_block_label <- function(label, fit){
  if(label == "log(tau_sq)")
    return("tau_sq")

  terms <- fit$chains[[1]]$backend$process_terms

  for(term in terms){
    sigma_label <- paste0("log(", term$name, "_sigma_sq)")
    if(label == sigma_label)
      return(paste0(term$name, "_sigma_sq"))

    prefix <- paste0(term$name, "_theta_")
    if(startsWith(label, prefix)){
      theta_index <- as.integer(substring(label, nchar(prefix) + 1L))
      theta_name <- term$theta_names[theta_index]
      return(paste0(term$name, "_", theta_name))
    }
  }

  label
}

block_summary <- function(fit, label){
  blocks <- fit$chains[[1]]$adaptive_metropolis$blocks
  out <- data.frame()

  for(j in seq_along(blocks)){
    out <- rbind(
      out,
      data.frame(
        Blocking = label,
        Block = j,
        Dimension = blocks[[j]]$dimension,
        Acceptance = round(blocks[[j]]$acceptance, 3),
        Parameters = paste(
          vapply(blocks[[j]]$parameter_labels, display_block_label, character(1), fit = fit),
          collapse = ", "
        ),
        check.names = FALSE
      )
    )
  }

  out
}

show_table(rbind(
  block_summary(joint_run, "joint"),
  block_summary(by_term_run, "by_term")
))
```

Next use multi-chain diagnostics. `summary()` for a multi-chain `stLMM` fit reports diagnostics computed with `coda`, and `as_mcmc()` returns a `coda::mcmc.list` if you want to work directly with the chains. The `burn` and `thin` arguments can be passed to `as_mcmc()` directly, including for recovered multi-chain objects. Here we focus on $\hat R$, effective sample size (ESS), and ESS per second for the Metropolis-updated parameters.

```{r coda-diagnostics}
#| echo: false
covariance_parameters <- c("tau_sq", "nngp_1_sigma_sq", "nngp_1_phi")

coda_diagnostics <- function(run, label){
  chains <- as_mcmc(run)
  chains <- chains[, covariance_parameters]

  rhat <- coda::gelman.diag(chains, multivariate = FALSE)$psrf[, "Point est."]
  ess <- coda::effectiveSize(chains)
  elapsed <- run$timing$sampler_total[["elapsed"]]

  data.frame(
    Blocking = label,
    Parameter = covariance_parameters,
    Rhat = round(as.numeric(rhat[covariance_parameters]), 3),
    ESS = round(as.numeric(ess[covariance_parameters]), 1),
    `ESS per second` = round(as.numeric(ess[covariance_parameters]) / elapsed, 1),
    check.names = FALSE
  )
}

show_table(rbind(
  coda_diagnostics(joint_run, "joint"),
  coda_diagnostics(by_term_run, "by_term")
))
```

It's also helpful to summarize the cost of each blocking scheme. A blocked sampler can have better convergence diagnostics per retained iteration but still lose on effective sample size per second if the extra blocks make each iteration much slower.

```{r runtime-summary}
#| echo: false
runtime_summary <- function(run, label){
  diag <- coda_diagnostics(run, label)

  data.frame(
    Blocking = label,
    `Sampler elapsed seconds` = round(run$timing$sampler_total[["elapsed"]], 2),
    `Worst Rhat` = max(diag$Rhat, na.rm = TRUE),
    `Minimum ESS` = min(diag$ESS, na.rm = TRUE),
    `Minimum ESS per second` = min(diag$`ESS per second`, na.rm = TRUE),
    check.names = FALSE
  )
}

show_table(rbind(
  runtime_summary(joint_run, "joint"),
  runtime_summary(by_term_run, "by_term")
))
```

In this example, the default joint block is the baseline: it keeps the process variance and decay parameter in the same proposal and avoids extra target evaluations. The `"by_term"` option asks a specific question: does the residual variance need to move separately from the NNGP covariance parameters? In this run, the answer is no. The added block costs more computation and doesn't improve the chain diagnostics enough to beat the default joint update. That is the teaching point. Blocking can be useful, but it has to earn its cost in better mixing, higher ESS, or better ESS per second.

Trace plots are still worth looking at. $\hat R$ and ESS can miss local behavior, especially in short runs.

```{r trace-plots-joint, fig.width=8, fig.height=5}
#| code-fold: true
#| code-summary: "Show joint run plotting code"
chain_colors <- stlmm_discrete_colors(3)

plot(
  joint_run,
  parameters = covariance_parameters,
  n_col = 2,
  chain_colors = chain_colors
)
```

```{r trace-plots-by-term, fig.width=8, fig.height=5}
#| code-fold: true
#| code-summary: "Show by_term run plotting code"
plot(
  by_term_run,
  parameters = covariance_parameters,
  n_col = 2,
  chain_colors = chain_colors
)
```

## Changing Adaptation Settings

The previous fits used the default retained target acceptance for multivariate blocks. You can change retained adaptation through the `metropolis` list. The call below uses `"variance_theta"` blocking, longer retained adaptation batches, and a slightly higher target acceptance rate.

```{r custom-controls, eval=FALSE}
fit_custom <- stLMM(
  y ~ x + nngp(x_coord, y_coord, m = 15,
               cov_model = "exp", ordering = "maxmin"),
  data = dat,
  priors = list(
    resid = list(tau_sq = half_t(df = 3, scale = 0.5)),
    nngp_1 = list(
      sigma_sq = half_t(df = 3, scale = 1),
      phi = uniform(1, 15)
    )
  ),
  warmup = list(
    batch_length = 50,
    min_batches = 3,
    max_batches = 30,
    target = c(0.20, 0.40)
  ),
  metropolis = list(
    blocking = "variance_theta",
    batch_length = 50,
    target_accept = 0.30
  )
)
```

The scalar option is available too.

```{r scalar-controls, eval=FALSE}
fit_scalar <- stLMM(
  y ~ x + nngp(x_coord, y_coord, m = 15,
               cov_model = "exp", ordering = "maxmin"),
  data = dat,
  priors = list(
    resid = list(tau_sq = half_t(df = 3, scale = 0.5)),
    nngp_1 = list(
      sigma_sq = half_t(df = 3, scale = 1),
      phi = uniform(1, 15)
    )
  ),
  metropolis = "scalar"
)
```

Scalar updates can help diagnose whether a multivariate block is too hard to tune. They shouldn't be treated as a routine default for collapsed process models because each scalar proposal can require another collapsed target evaluation and factorization of $\mathbf{M}$.

## Practical Guidance

Start with `"joint"`. It's the default because it's usually fastest and often mixes well enough.

Try `"variance_theta"` when variance-like parameters and bounded range or correlation parameters appear to move on different scales. This is a common first alternative for a single structured process term.

Try `"by_term"` when a model has several structured process terms and one term appears to tune differently from the others.

Try `"process_variance"` when residual variance, process variance, and range or correlation parameters each need separate movement.

Try `"residual_process"` when residual variances dominate the joint proposal.

Use `"scalar"` as a diagnostic fallback. It can improve movement for stubborn parameters, but it can be much slower because it may multiply the number of collapsed target evaluations per retained iteration.

Compare blocking choices with several criteria: trace plots, block acceptance rates, $\hat R$, effective sample size, effective sample size per second, posterior summaries, and runtime. Better mixing per iteration only helps if it improves mixing enough to justify the extra target evaluations.

For difficult models, treat blocking as one diagnostic tool among several rather than as evidence that one parameterization is uniformly better.
