---
title: "Improving stLMM runtime on multicore systems"
subtitle: "Initial practical guidance for OpenMP, BLAS/LAPACK, and CHOLMOD"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: show
    code-overflow: wrap
    html-math-method: katex
execute:
  warning: false
  message: false
bibliography: references.bib
---

## Purpose

This vignette gives practical starting points for improving `stLMM` runtime on
multicore systems. The recommendations are intentionally cautious. Runtime
depends on the model, data structure, sparse-matrix fill-in, hardware, memory
bandwidth, BLAS/LAPACK implementation, and the `Matrix`/CHOLMOD build used by R.

More testing is needed across machines and model classes. Treat these notes as
initial guidance rather than fixed rules. Users are encouraged to run small
timing experiments on their own analyses, and feedback about configurations that
help or hurt is very welcome.

## Runtime Layers

There are three computational layers to keep separate.

First, `stLMM` has OpenMP-parallel sections in selected compiled kernels. These
are controlled by the `n_omp_threads` argument. This is not a general "make
everything parallel" switch; it only applies to code regions that were written
with OpenMP.

Second, dense linear algebra is handled by BLAS and LAPACK
[@BlackfordEtAl2002BLAS; @AndersonEtAl1999LAPACK]. R supplies these libraries to
packages. If R is linked against an optimized implementation such as OpenBLAS
[@OpenBLAS2025], MKL, or BLIS, dense linear algebra can be faster. If that
implementation is threaded, BLAS/LAPACK can also create its own worker threads.

Third, sparse Cholesky factorization is handled by CHOLMOD
[@ChenDavisHagerRajamanickam2008CHOLMOD] through the R `Matrix` package
[@BatesMaechlerJagan2025Matrix]. This layer is important for NNGP, CAR,
CAR-time, and autoregressive terms because these terms produce sparse precision
matrices. Sparse Cholesky runtime depends strongly on ordering, fill-in,
supernode sizes, symbolic structure, and memory traffic.

These layers have different controls. `n_omp_threads` does not set OpenBLAS,
MKL, or BLIS threads. BLAS thread controls do not necessarily control CHOLMOD's
own OpenMP threading. Increasing several thread counts at once can oversubscribe
the machine and make a run slower.

## Practical Observations

During development benchmarks, the most reliable improvement came from
`stLMM`'s own OpenMP regions: runtime improved as `n_omp_threads` increased up
to a point, then flattened or worsened. The useful number of threads depended on
the machine and workload.

Threaded BLAS did not improve the benchmark workloads we tested. In those runs,
single-threaded BLAS combined with `stLMM` OpenMP was fastest. When a verified
threaded BLAS was used, CHOLMOD-related time became slower, and prediction-heavy
timings did not improve. This is an observation from specific development
benchmarks, not a guarantee for every model or machine.

CHOLMOD threading needs special care. Many R `Matrix` installations use CHOLMOD
without CHOLMOD OpenMP support. In that case, CHOLMOD's own OpenMP controls
cannot help. Supernodal CHOLMOD can still call BLAS for dense frontal-matrix
operations, so a threaded BLAS may be used inside CHOLMOD, but whether this helps
depends on the sparse structure, fill-in, and size of dense updates. For some
workloads, threaded BLAS overhead can outweigh any benefit.

The practical starting point is therefore:

- start with BLAS threads set to one;
- sweep `n_omp_threads` over a small grid appropriate for the machine;
- only then test BLAS threads if dense linear algebra is suspected to matter;
- change one layer at a time and keep the model, data, seed, and sample count
  fixed.

## Checking the Current Setup

Before timing, record the R and numerical-library setup.

```{r setup-info, eval=FALSE}
sessionInfo()
extSoftVersion()["BLAS"]
packageVersion("Matrix")
packageVersion("stLMM")
```

Do not assume that an OpenBLAS library is threaded just because its path contains
`openblas`. Some systems provide both single-threaded and threaded OpenBLAS
builds. On Linux, `ldd` can help show which libraries R and packages load.

```sh
ldd "$(R RHOME)/lib/libR.so"
ldd "$(R RHOME)/lib/libRblas.so"

Rscript -e 'so <- list.files(system.file("libs", package = "Matrix"), full.names = TRUE); system2("ldd", so)'
Rscript -e 'so <- list.files(system.file("libs", package = "stLMM"), full.names = TRUE); system2("ldd", so)'
```

If you are testing threaded BLAS, verify actual thread use rather than relying
only on environment variables. On Linux, one simple check during a running
benchmark is:

```sh
ps -o pid,nlwp,pcpu,comm,args -p <R-process-id>
```

`NLWP` is the number of lightweight process threads. It should increase when a
threaded BLAS or OpenMP region is active.

## Thread Controls

For interactive work, `RhpcBLASctl` is often convenient if the linked BLAS
supports runtime control.

```{r rhpcblasctl-example, eval=FALSE}
library(RhpcBLASctl)

RhpcBLASctl::blas_set_num_threads(1)
```

For reproducible benchmarks, prefer setting thread counts before R starts. For
example, with OpenBLAS:

```sh
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript timing-script.R
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=4 Rscript timing-script.R
OPENBLAS_NUM_THREADS=4 OMP_NUM_THREADS=1 Rscript timing-script.R
```

For `stLMM`'s own OpenMP regions, use the explicit model argument.

```{r omp-example, eval=FALSE}
fit <- stLMM(
  y ~ x + nngp(lon, lat, m = 15),
  data = dat,
  n_samples = 1000,
  n_omp_threads = 4,
  priors = list(
    resid = list(tau_sq = ig(2, 1)),
    nngp_1 = list(sigma_sq = ig(2, 1), phi = uniform(0.1, 10))
  )
)
```

Choose a thread grid that fits the machine. A laptop, a shared server, and a
large workstation should not use the same grid. A reasonable first sweep is a
small set of powers of two, capped at a sensible value for the machine.

```{r thread-grid-example, eval=FALSE}
physical_cores <- parallel::detectCores(logical = FALSE)
if (is.na(physical_cores)) {
  physical_cores <- parallel::detectCores(logical = TRUE)
}

omp_grid <- unique(pmin(c(1, 2, 4, 8, 16, physical_cores), physical_cores))
omp_grid
```

On larger machines, it can also be useful to include the number of logical CPUs.
For compute-heavy workloads, logical CPUs from simultaneous multithreading do
not usually double throughput, but they are worth testing if the run is memory
or latency limited.

## Matrix, CHOLMOD, and Ordering

`stLMM` uses the CHOLMOD interface exposed by `Matrix`. Installing SuiteSparse
or METIS system packages does not necessarily change an already installed
`Matrix` package. METIS-based orderings are available only if the CHOLMOD
library used by `Matrix` was built with that support.

The `cholmod_control` argument can be passed as a string, for example
`cholmod_control = "metis"`, or as a control list with an `ordering` entry. The
supported ordering values are `"auto"`, `"best"`, `"natural"`, `"amd"`,
`"metis"`, `"nesdis"`, and `"colamd"`. The default `"auto"` uses CHOLMOD's
default strategy. The `"best"` option asks CHOLMOD to try its full built-in
method set, which can make symbolic analysis slower.

For large sparse models, ordering is worth testing. The best ordering can differ
by model structure, spatial/temporal support, neighbor count, and Matrix/CHOLMOD
build.

```{r cholmod-ordering-example, eval=FALSE}
fit <- stLMM(
  y ~ x + nngp(lon, lat, m = 15),
  data = dat,
  n_samples = 1000,
  n_omp_threads = 4,
  cholmod_control = "amd",
  priors = list(
    resid = list(tau_sq = ig(2, 1)),
    nngp_1 = list(sigma_sq = ig(2, 1), phi = uniform(0.1, 10))
  )
)
```

## Benchmarking Your Own Runs

The most useful benchmark is the target analysis, or a shortened version of it.
Keep the model, data, seed, and number of MCMC samples fixed while changing one
runtime setting at a time.

A first experiment is to hold BLAS threads at one and vary `n_omp_threads`.

```r
run_fit <- function(omp_threads) {
  stLMM(
    y ~ x + nngp(lon, lat, m = 15),
    data = dat,
    n_samples = 200,
    n_omp_threads = omp_threads,
    priors = list(
      resid = list(tau_sq = ig(2, 1)),
      nngp_1 = list(sigma_sq = ig(2, 1), phi = uniform(0.1, 10))
    )
  )
}

system.time(run_fit(1))
system.time(run_fit(4))
system.time(run_fit(8))
```

For more reproducible comparisons, use separate R processes so thread
environment variables are fixed before R starts.

```sh
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=1 Rscript timing-script.R
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=4 Rscript timing-script.R
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=8 Rscript timing-script.R
```

Then, if you want to test threaded BLAS, hold `n_omp_threads` fixed and vary the
BLAS layer. Monitor actual thread use and elapsed time.

```sh
OPENBLAS_NUM_THREADS=1 OMP_NUM_THREADS=4 Rscript timing-script.R
OPENBLAS_NUM_THREADS=2 OMP_NUM_THREADS=4 Rscript timing-script.R
OPENBLAS_NUM_THREADS=4 OMP_NUM_THREADS=4 Rscript timing-script.R
```

Before sharing benchmark results, record:

- CPU model, number of physical cores, number of logical CPUs, and memory;
- `sessionInfo()` and `extSoftVersion()["BLAS"]`;
- `Matrix` and `stLMM` versions;
- loaded BLAS/LAPACK libraries, when available;
- thread settings such as `OPENBLAS_NUM_THREADS`, `MKL_NUM_THREADS`,
  `BLIS_NUM_THREADS`, and `OMP_NUM_THREADS`;
- model family, sparse term type, neighbor count, data size, and
  `cholmod_control`;
- elapsed time summaries across repeated runs when possible.

## Practical Checklist

- Use an optimized BLAS/LAPACK implementation, but do not assume threaded BLAS
  will help.
- Start benchmarking with BLAS threads set to one.
- Sweep `n_omp_threads` over a machine-appropriate grid.
- Avoid setting BLAS and OpenMP thread counts high at the same time unless
  oversubscription is the experiment.
- For large sparse models, test CHOLMOD ordering choices.
- Verify actual thread use when testing threaded libraries.
- Repeat timing experiments for each major model/data class.
- Share feedback when alternative BLAS/LAPACK, Matrix/CHOLMOD builds, ordering
  choices, or thread settings materially improve performance.
