---
title: "Independent random effects: stLMM and lme4"
subtitle: "Gaussian, logistic, Poisson-approximation, and negative-binomial comparisons"
author: "Andrew O. Finley"
bibliography: references.bib
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    code-overflow: wrap
    html-math-method: katex
    css: stlmm-vignette.css
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
Sys.setenv(
  OPENBLAS_NUM_THREADS = "1",
  MKL_NUM_THREADS = "1",
  BLIS_NUM_THREADS = "1",
  VECLIB_MAXIMUM_THREADS = "1"
)

if(requireNamespace("RhpcBLASctl", quietly = TRUE))
  RhpcBLASctl::blas_set_num_threads(1)

library(stLMM)
library(lme4)
library(ggplot2)

vignette_dir <- dirname(knitr::current_input(dir = TRUE))
source(file.path(vignette_dir, "utils.R"))

set.seed(1)
theme_set(theme_bw(base_size = 12))
```

## Purpose

This article compares `stLMM` with `lme4` for models with independent grouped
random effects. The goal is comparison, not replacement. `lme4` is an expansive,
mature, and widely used mixed-model package. @BatesMaechlerBolkerWalker2015lme4
describe `lme4` as a modular system for fitting linear and generalized linear
mixed models using sparse-matrix methods and profiled likelihood or REML
estimation.

There are several important things `lme4` can do that `stLMM` currently cannot:

- estimate covariance among random-effect terms, as in `(1 + x | group)`;
- fit nested and crossed random-effect structures through the standard formula
  interface;
- fit models by ML or REML and report likelihood-based diagnostics;
- support a broad GLMM workflow through `glmer()`.

The fair comparison is narrower. We compare models with fixed effects, one or
more independent group-level random effects, and likelihoods that both packages
can represent exactly or approximately. In `stLMM`, a random intercept and two
independent random slopes are written as

```r
y ~ x1 + x2 + iid(group) + x1:iid(group) + x2:iid(group)
```

The matching `lme4` formula is

```r
y ~ x1 + x2 + (1 | group) + (0 + x1 | group) + (0 + x2 | group)
```

For numeric random-effect terms, this is also equivalent to the compact `lme4`
formula

```r
y ~ x1 + x2 + (1 + x1 + x2 || group)
```

This intentionally avoids `(1 + x1 + x2 | group)`, because that model estimates
the covariance matrix among group-level intercepts and slopes.

The non-Gaussian comparisons need extra caution:

- `stLMM` uses Pólya-Gamma data augmentation for binomial and
  negative-binomial likelihoods;
- `lme4::glmer()` fits GLMMs by likelihood approximation;
- `stLMM` represents Poisson models through a large-size negative-binomial
  approximation, not a native Poisson likelihood;
- `stLMM` currently fixes the negative-binomial size parameter, while
  `lme4::glmer.nb()` estimates it.

```{r helpers}
#| code-fold: true
#| code-summary: "Show helper functions"
posterior_means <- function(fit, burn = 500L){
  draws <- as_samples(fit, burn = burn, metadata = FALSE)
  draws$.chain <- NULL
  draws$.iteration <- NULL
  vapply(draws, mean, numeric(1))
}

posterior_sds <- function(fit, burn = 500L){
  draws <- as_samples(fit, burn = burn, metadata = FALSE)
  draws$.chain <- NULL
  draws$.iteration <- NULL
  vapply(draws, sd, numeric(1))
}

round_numeric_columns <- function(x, digits = 3){
  is_num <- vapply(x, is.numeric, logical(1))
  x[is_num] <- lapply(x[is_num], round, digits = digits)
  x
}

format_estimate_with_uncertainty <- function(estimate, uncertainty, digits = 3){
  paste0(
    format(round(estimate, digits), nsmall = digits),
    " (",
    format(round(uncertainty, digits), nsmall = digits),
    ")"
  )
}

fixed_effect_table <- function(st_fit, lme4_fit, truth, burn = 500L){
  st_mean <- posterior_means(st_fit, burn = burn)
  st_sd <- posterior_sds(st_fit, burn = burn)
  lme4_beta <- fixef(lme4_fit)
  lme4_se <- sqrt(diag(vcov(lme4_fit)))

  data.frame(
    parameter = names(lme4_beta),
    truth = unname(truth[names(lme4_beta)]),
    stLMM = format_estimate_with_uncertainty(
      unname(st_mean[names(lme4_beta)]),
      unname(st_sd[names(lme4_beta)])
    ),
    lme4 = format_estimate_with_uncertainty(
      unname(lme4_beta),
      unname(lme4_se[names(lme4_beta)])
    ),
    difference = unname(st_mean[names(lme4_beta)] - lme4_beta),
    row.names = NULL
  )
}

diagnostic_table <- function(st_fit, parameters, burn = 500L){
  s <- summary(st_fit, burn = burn)
  s$diagnostics[parameters, c("parameter", "rhat", "effective_size"),
                drop = FALSE]
}

iid_random_effect_compare <- function(st_fit, lme4_fit, st_term = 1L,
                                      lme4_column = "(Intercept)",
                                      burn = 500L){
  st_mean <- posterior_means(st_fit, burn = burn)
  prefix <- paste0("iid_", st_term, "_")
  st_re <- st_mean[grep(paste0("^", prefix), names(st_mean))]
  names(st_re) <- sub(paste0("^", prefix), "", names(st_re))

  lme4_re <- ranef(lme4_fit)$group[[lme4_column]]
  names(lme4_re) <- rownames(ranef(lme4_fit)$group)

  common <- intersect(names(st_re), names(lme4_re))
  data.frame(
    group = common,
    stLMM = unname(st_re[common]),
    lme4 = unname(lme4_re[common])
  )
}
```

## Gaussian Models

Gaussian mixed models are the cleanest comparison because both packages target
the same linear mixed-model likelihood for independent random effects.

### Simple Data

Each group has its own intercept and its own slope for `x`, and those two
group-level effects are independent in the data-generating model.

```{r simulate-gaussian-small}
#| code-fold: true
#| code-summary: "Show simulation code"
simulate_gaussian_slope_data <- function(seed = 1,
                                         n_group = 30,
                                         n_per_group = 15,
                                         beta = c("(Intercept)" = 1, x = 2),
                                         sigma_intercept = 0.5,
                                         sigma_slope = 0.5,
                                         tau = 0.5){
  set.seed(seed)

  group <- factor(rep(sprintf("g%02d", seq_len(n_group)),
                      each = n_per_group))
  x <- rep(seq(-1, 1, length.out = n_per_group), n_group) +
    rnorm(n_group * n_per_group, sd = 0.08)

  alpha_intercept <- rnorm(n_group, sd = sigma_intercept)
  alpha_slope <- rnorm(n_group, sd = sigma_slope)
  names(alpha_intercept) <- levels(group)
  names(alpha_slope) <- levels(group)

  mu <- beta[["(Intercept)"]] + beta[["x"]] * x +
    alpha_intercept[group] + alpha_slope[group] * x

  data.frame(
    y = mu + rnorm(length(mu), sd = tau),
    x = x,
    group = group
  )
}

gauss_dat <- simulate_gaussian_slope_data(seed = 1)
gauss_true_beta <- c("(Intercept)" = 1, x = 2)
gauss_true_sd <- c(
  residual = 0.5,
  `random intercept` = 0.5,
  `random slope x` = 0.5
)
```

The simulated dataset has 30 groups with 15 observations in each group.

```{r gaussian-data-plot, fig.width=7, fig.height=4}
#| code-fold: true
#| code-summary: "Show plotting code"
ggplot(gauss_dat, aes(x, y, group = group)) +
  geom_line(aes(color = group), alpha = 0.3, linewidth = 0.35) +
  geom_point(aes(color = group), alpha = 0.75, size = 1.2) +
  scale_color_manual(
    values = rep(stlmm_discrete_colors(8),
                 length.out = nlevels(gauss_dat$group))
  ) +
  guides(color = "none") +
  labs(x = "x", y = "response")
```

### Fits

The `stLMM` fit uses weak inverse-gamma priors for the Gaussian residual
variance and the two independent iid random-effect variances. Posterior
inference is based on three chains, with $\hat R$ diagnostics reported below.
The `lme4` fit uses REML.

```{r fit-gaussian-small}
gauss_st_time <- system.time({
  gauss_st <- stLMM(
    y ~ x + iid(group) + x:iid(group),
    data = gauss_dat,
    priors = list(
      resid = list(tau_sq = ig(2, 0.5)),
      iid_1 = list(sigma_sq = ig(2, 0.5)),
      iid_2 = list(sigma_sq = ig(2, 0.5))
    ),
    n_samples = 1500,
    chains = 3,
    verbose = FALSE
  )
})

gauss_lme4_time <- system.time({
  gauss_lme4 <- lmer(
    y ~ x + (1 | group) + (0 + x | group),
    data = gauss_dat,
    REML = TRUE
  )
})
```

### Parameter Estimates

Fixed-effect values are shown as estimate followed by uncertainty in
parentheses. For `stLMM`, the parenthetical value is posterior SD; for `lme4`,
it is the fixed-effect SE reported by `summary()`.

The random-effect standard-deviation table reports point estimates only. Those
variance-component summaries are useful for checking scale, but they should not
be read as equally precise across packages.

```{r gaussian-small-tables}
#| code-fold: true
#| code-summary: "Show table code"
gauss_st_mean <- posterior_means(gauss_st, burn = 500)
gauss_lme4_vc <- as.data.frame(VarCorr(gauss_lme4))
gauss_lme4_vc <- gauss_lme4_vc[gauss_lme4_vc$grp != "Residual", ,
                               drop = FALSE]

gauss_beta_tab <- fixed_effect_table(
  gauss_st, gauss_lme4, truth = gauss_true_beta, burn = 500
)
gauss_sd_tab <- data.frame(
  component = names(gauss_true_sd),
  truth = unname(gauss_true_sd),
  stLMM = c(
    sqrt(gauss_st_mean[["tau_sq"]]),
    sqrt(gauss_st_mean[["iid_1_sigma_sq"]]),
    sqrt(gauss_st_mean[["iid_2_sigma_sq"]])
  ),
  lme4 = c(sigma(gauss_lme4), gauss_lme4_vc$sdcor),
  row.names = NULL
)
gauss_sd_tab$difference <- gauss_sd_tab$stLMM - gauss_sd_tab$lme4
gauss_diag <- diagnostic_table(
  gauss_st,
  parameters = c("(Intercept)", "x", "tau_sq",
                 "iid_1_sigma_sq", "iid_2_sigma_sq"),
  burn = 500
)

knitr::kable(round_numeric_columns(gauss_beta_tab, 3))
knitr::kable(round_numeric_columns(gauss_sd_tab, 3))
knitr::kable(round_numeric_columns(gauss_diag, 3))
```

```{r gaussian-random-effect-plot, fig.width=7, fig.height=4}
#| code-fold: true
#| code-summary: "Show plotting code"
gauss_re <- rbind(
  cbind(term = "intercept",
        iid_random_effect_compare(gauss_st, gauss_lme4, st_term = 1,
                                  lme4_column = "(Intercept)", burn = 500)),
  cbind(term = "slope x",
        iid_random_effect_compare(gauss_st, gauss_lme4, st_term = 2,
                                  lme4_column = "x", burn = 500))
)

ggplot(gauss_re, aes(lme4, stLMM)) +
  geom_abline(slope = 1, intercept = 0, color = "grey45", linewidth = 0.4) +
  geom_point(size = 1.8, alpha = 0.85, color = stlmm_color("primary")) +
  facet_wrap(~ term, scales = "free") +
  labs(
    x = "lme4 conditional mode",
    y = "stLMM posterior mean"
  )
```

## Logistic Mixed Model

The logistic comparison uses Bernoulli responses with a logit link and an
independent group random intercept.

```{r simulate-logistic}
#| code-fold: true
#| code-summary: "Show logistic simulation code"
simulate_logistic_data <- function(seed = 1,
                                   n_group = 30,
                                   n_per_group = 12,
                                   beta = c("(Intercept)" = -0.5, x = 1),
                                   sigma_intercept = 0.5){
  set.seed(seed)

  group <- factor(rep(sprintf("g%02d", seq_len(n_group)),
                      each = n_per_group))
  x <- rnorm(n_group * n_per_group)
  alpha <- rnorm(n_group, sd = sigma_intercept)
  names(alpha) <- levels(group)
  eta <- beta[["(Intercept)"]] + beta[["x"]] * x + alpha[group]

  data.frame(
    y = rbinom(length(eta), size = 1, prob = plogis(eta)),
    x = x,
    group = group
  )
}

logit_dat <- simulate_logistic_data(seed = 1)
logit_truth <- c("(Intercept)" = -0.5, x = 1)
logit_sigma <- 0.5
```

```{r fit-logistic}
logit_st_time <- system.time({
  logit_st <- stLMM(
    y ~ x + iid(group),
    data = logit_dat,
    family = "binomial",
    priors = list(iid_1 = list(sigma_sq = ig(2, 0.5))),
    n_samples = 1000,
    chains = 3,
    verbose = FALSE
  )
})

logit_lme4_time <- system.time({
  logit_lme4 <- glmer(
    y ~ x + (1 | group),
    data = logit_dat,
    family = binomial,
    nAGQ = 1
  )
})
```

```{r logistic-tables}
#| code-fold: true
#| code-summary: "Show logistic table code"
logit_st_mean <- posterior_means(logit_st, burn = 500)
logit_vc <- as.data.frame(VarCorr(logit_lme4))
logit_vc <- logit_vc[logit_vc$grp != "Residual", , drop = FALSE]
logit_beta_tab <- fixed_effect_table(
  logit_st, logit_lme4, truth = logit_truth, burn = 500
)
logit_sd_tab <- data.frame(
  component = "random intercept",
  truth = logit_sigma,
  stLMM = sqrt(logit_st_mean[["iid_1_sigma_sq"]]),
  lme4 = logit_vc$sdcor[1],
  difference = sqrt(logit_st_mean[["iid_1_sigma_sq"]]) - logit_vc$sdcor[1]
)
logit_diag <- diagnostic_table(
  logit_st,
  parameters = c("(Intercept)", "x", "iid_1_sigma_sq"),
  burn = 500
)

knitr::kable(round_numeric_columns(logit_beta_tab, 3))
knitr::kable(round_numeric_columns(logit_sd_tab, 3))
knitr::kable(round_numeric_columns(logit_diag, 3))
```

In this simulation, the fixed-effect estimates are close across packages. The
random-intercept standard deviation is closer to the simulated truth under
`stLMM` than under `lme4`, while the group-level conditional modes still align
closely in the plot below. This illustrates a small finite-sample difference
between the two estimation approaches.

```{r logistic-random-effect-plot, fig.width=5.5, fig.height=4}
#| code-fold: true
#| code-summary: "Show plotting code"
logit_re <- iid_random_effect_compare(logit_st, logit_lme4, burn = 500)

ggplot(logit_re, aes(lme4, stLMM)) +
  geom_abline(slope = 1, intercept = 0, color = "grey45", linewidth = 0.4) +
  geom_point(size = 1.8, alpha = 0.85, color = stlmm_color("primary")) +
  labs(
    x = "lme4 conditional mode",
    y = "stLMM posterior mean"
  )
```

## Poisson Data and the NB Approximation

`stLMM` does not currently fit a native Poisson likelihood. The package can fit
a negative-binomial model with large fixed size, where

$$
\operatorname{var}(y_i) = \mu_i + \frac{\mu_i^2}{r},
$$

and this approaches the Poisson variance as the size $r$ becomes large. This is
an approximation check, not an exact likelihood comparison. We use
`size = 1000`, matching the large-size setting used in the Scottish lip cancer
comparison article. This large size can mix more slowly, so this section uses
longer chains than the logistic example.

```{r simulate-poisson}
#| code-fold: true
#| code-summary: "Show Poisson simulation code"
simulate_poisson_data <- function(seed = 1,
                                  n_group = 30,
                                  n_per_group = 12,
                                  beta = c("(Intercept)" = 0.5, x = 0.5),
                                  sigma_intercept = 0.5){
  set.seed(seed)

  group <- factor(rep(sprintf("g%02d", seq_len(n_group)),
                      each = n_per_group))
  x <- rnorm(n_group * n_per_group)
  alpha <- rnorm(n_group, sd = sigma_intercept)
  names(alpha) <- levels(group)
  eta <- beta[["(Intercept)"]] + beta[["x"]] * x + alpha[group]

  data.frame(
    y = rpois(length(eta), lambda = exp(eta)),
    x = x,
    group = group
  )
}

poisson_dat <- simulate_poisson_data(seed = 1)
poisson_truth <- c("(Intercept)" = 0.5, x = 0.5)
poisson_sigma <- 0.5
poisson_approx_size <- 1000
```

```{r fit-poisson}
poisson_st_time <- system.time({
  poisson_st <- stLMM(
    y ~ x + iid(group),
    data = poisson_dat,
    family = "negative_binomial",
    size = poisson_approx_size,
    priors = list(iid_1 = list(sigma_sq = ig(2, 0.5))),
    n_samples = 5000,
    chains = 3,
    verbose = FALSE
  )
})

poisson_lme4_time <- system.time({
  poisson_lme4 <- glmer(
    y ~ x + (1 | group),
    data = poisson_dat,
    family = poisson
  )
})
```

```{r poisson-tables}
#| code-fold: true
#| code-summary: "Show Poisson-approximation table code"
poisson_st_mean <- posterior_means(poisson_st, burn = 2500)
poisson_vc <- as.data.frame(VarCorr(poisson_lme4))
poisson_vc <- poisson_vc[poisson_vc$grp != "Residual", , drop = FALSE]
poisson_beta_tab <- fixed_effect_table(
  poisson_st, poisson_lme4, truth = poisson_truth, burn = 2500
)
poisson_sd_tab <- data.frame(
  component = "random intercept",
  truth = poisson_sigma,
  stLMM = sqrt(poisson_st_mean[["iid_1_sigma_sq"]]),
  lme4 = poisson_vc$sdcor[1],
  difference = sqrt(poisson_st_mean[["iid_1_sigma_sq"]]) - poisson_vc$sdcor[1]
)
poisson_diag <- diagnostic_table(
  poisson_st,
  parameters = c("(Intercept)", "x", "iid_1_sigma_sq"),
  burn = 2500
)

knitr::kable(round_numeric_columns(poisson_beta_tab, 3))
knitr::kable(round_numeric_columns(poisson_sd_tab, 3))
knitr::kable(round_numeric_columns(poisson_diag, 3))
```

The Poisson approximation recovers the fixed effects reasonably well. The
diagnostics also show that this large-size NB approximation gives a more
challenging posterior computation than the logistic model. The intercept
$\hat R$ remains above the usual target even after the longer run, so this
section should be read as both an estimator comparison and a caution about using
the current approximation as a stand-in for a native Poisson likelihood.

```{r poisson-random-effect-plot, fig.width=5.5, fig.height=4}
#| code-fold: true
#| code-summary: "Show plotting code"
poisson_re <- iid_random_effect_compare(poisson_st, poisson_lme4, burn = 2500)

ggplot(poisson_re, aes(lme4, stLMM)) +
  geom_abline(slope = 1, intercept = 0, color = "grey45", linewidth = 0.4) +
  geom_point(size = 1.8, alpha = 0.85, color = stlmm_color("primary")) +
  labs(
    x = "lme4 conditional mode",
    y = "stLMM posterior mean"
  )
```

## Negative-Binomial Mixed Model

The final example simulates negative-binomial counts with fixed size 3. The
`stLMM` fit uses this known fixed size. The `lme4::glmer.nb()` fit estimates the
size parameter, so this is close but not exactly the same estimation problem.
Estimating the negative-binomial size parameter in `stLMM` would make this
comparison cleaner and is a natural future extension.

```{r simulate-nb}
#| code-fold: true
#| code-summary: "Show negative-binomial simulation code"
simulate_nb_data <- function(seed = 1,
                             n_group = 40,
                             n_per_group = 12,
                             beta = c("(Intercept)" = 0.5, x = 0.5),
                             sigma_intercept = 0.5,
                             size = 3){
  set.seed(seed)

  group <- factor(rep(sprintf("g%02d", seq_len(n_group)),
                      each = n_per_group))
  x <- rnorm(n_group * n_per_group)
  alpha <- rnorm(n_group, sd = sigma_intercept)
  names(alpha) <- levels(group)
  eta <- beta[["(Intercept)"]] + beta[["x"]] * x + alpha[group]

  data.frame(
    y = rnbinom(length(eta), size = size, mu = exp(eta)),
    x = x,
    group = group
  )
}

nb_size <- 3
nb_dat <- simulate_nb_data(seed = 1, size = nb_size)
nb_truth <- c("(Intercept)" = 0.5, x = 0.5)
nb_sigma <- 0.5
```

```{r fit-nb}
nb_st_time <- system.time({
  nb_st <- stLMM(
    y ~ x + iid(group),
    data = nb_dat,
    family = "negative_binomial",
    size = nb_size,
    priors = list(iid_1 = list(sigma_sq = ig(2, 0.5))),
    n_samples = 1000,
    chains = 3,
    verbose = FALSE
  )
})

nb_lme4_time <- system.time({
  nb_lme4 <- glmer.nb(
    y ~ x + (1 | group),
    data = nb_dat
  )
})

nb_lme4_size <- getME(nb_lme4, "glmer.nb.theta")
```

```{r nb-tables}
#| code-fold: true
#| code-summary: "Show negative-binomial table code"
nb_st_mean <- posterior_means(nb_st, burn = 500)
nb_vc <- as.data.frame(VarCorr(nb_lme4))
nb_vc <- nb_vc[nb_vc$grp != "Residual", , drop = FALSE]
nb_beta_tab <- fixed_effect_table(
  nb_st, nb_lme4, truth = nb_truth, burn = 500
)
nb_sd_tab <- data.frame(
  component = "random intercept",
  truth = nb_sigma,
  stLMM = sqrt(nb_st_mean[["iid_1_sigma_sq"]]),
  lme4 = nb_vc$sdcor[1],
  difference = sqrt(nb_st_mean[["iid_1_sigma_sq"]]) - nb_vc$sdcor[1]
)
nb_size_tab <- data.frame(
  parameter = "negative-binomial size",
  truth = nb_size,
  stLMM = nb_size,
  lme4 = nb_lme4_size
)
nb_diag <- diagnostic_table(
  nb_st,
  parameters = c("(Intercept)", "x", "iid_1_sigma_sq"),
  burn = 500
)

knitr::kable(round_numeric_columns(nb_beta_tab, 3))
knitr::kable(round_numeric_columns(nb_sd_tab, 3))
knitr::kable(round_numeric_columns(nb_size_tab, 3))
knitr::kable(round_numeric_columns(nb_diag, 3))
```

The negative-binomial fixed effects and random-intercept scale are close across
the two packages, even though the comparison is not perfectly matched: `stLMM`
uses the known fixed size from the simulation, while `lme4::glmer.nb()` estimates
the size parameter from the data.

```{r nb-random-effect-plot, fig.width=5.5, fig.height=4}
#| code-fold: true
#| code-summary: "Show plotting code"
nb_re <- iid_random_effect_compare(nb_st, nb_lme4, burn = 500)

ggplot(nb_re, aes(lme4, stLMM)) +
  geom_abline(slope = 1, intercept = 0, color = "grey45", linewidth = 0.4) +
  geom_point(size = 1.8, alpha = 0.85, color = stlmm_color("primary")) +
  labs(
    x = "lme4 conditional mode",
    y = "stLMM posterior mean"
  )
```

## Runtime Summary

The runtime table records elapsed time for the fitted models above. The `stLMM`
timings are for posterior sampling, while the `lme4` timings are likelihood or
REML fits. These are different computational targets.

```{r runtime-table}
#| code-fold: true
#| code-summary: "Show runtime table code"
runtime_tab <- data.frame(
  example = c(
    "Gaussian simple", "Gaussian simple",
    "logistic", "logistic",
    "Poisson approximation", "Poisson approximation",
    "negative binomial", "negative binomial"
  ),
  package = rep(c("stLMM", "lme4"), times = 4),
  task = c(
    "3 chains x 1500 posterior draws", "REML fit",
    "3 chains x 1000 posterior draws", "glmer binomial",
    "3 chains x 5000 posterior draws", "glmer Poisson",
    "3 chains x 1000 posterior draws", "glmer.nb"
  ),
  elapsed_seconds = c(
    gauss_st_time[["elapsed"]],
    gauss_lme4_time[["elapsed"]],
    logit_st_time[["elapsed"]],
    logit_lme4_time[["elapsed"]],
    poisson_st_time[["elapsed"]],
    poisson_lme4_time[["elapsed"]],
    nb_st_time[["elapsed"]],
    nb_lme4_time[["elapsed"]]
  )
)

knitr::kable(round_numeric_columns(runtime_tab, 3))
```

## Takeaways

For Gaussian models with independent grouped random effects, `stLMM` and `lme4`
produce very similar fixed effects, variance components, and group-level
effects in this simulation.

The logistic comparison is the cleanest GLMM comparison. The Poisson example
shows that a large-size negative-binomial fit can track a native Poisson
`glmer()` fit for fixed effects, but it also shows that the approximation can
require more posterior simulation effort. The negative-binomial example is
useful, but the packages treat the size parameter differently: fixed in `stLMM`,
estimated in `lme4`. A native Poisson likelihood and an estimated
negative-binomial size parameter would both make future comparisons more direct.
