---
title: "Direct-estimate CAR-time model with scaled variances"
subtitle: "Area-response smoothing with recalibrated direct-estimate variances"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-overflow: wrap
    html-math-method: katex
    css: stlmm-vignette.css
execute:
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| code-summary: "Show setup code"
library(stLMM)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)

source("article-utils.R")

set.seed(1)
theme_set(theme_bw(base_size = 12))

article_id <- "wa-direct-car-time-scaled-variance"
data_dir <- Sys.getenv(
  "STLMM_WA_DATA_DIR",
  unset = file.path(article_helper_dir, "wa_data")
)
```

## Purpose

This article extends the direct-estimate CAR-time model by estimating a low-dimensional scaling model for the supplied direct-estimate variances. The response is a county-year direct estimate of live aboveground biomass density, and the inferential target remains the latent county-year biomass mean.

This is an **area-response** model: each row represents a county-year, not a plot. It follows the Fay-Herriot small-area model structure [@FayHerriot1979], with a CAR-time process replacing an independent area effect. Unlike the fixed-variance article, the supplied direct-estimate variances are treated as a variance template that the model can recalibrate.

Source code and data links for reproducing this article are collected below.

::: {.stlmm-resource-box .stlmm-resource-source}
<div class="stlmm-resource-title">Source Code</div>

- [Quarto source](wa-direct-car-time-scaled-variance.qmd)
- [Extracted R code](wa-direct-car-time-scaled-variance.R)
:::

::: {.stlmm-resource-box .stlmm-resource-data}
<div class="stlmm-resource-title">Data Bundle</div>

- [Direct-estimate table](wa_data/wa_direct_estimates.csv)
- [County geometry](wa_data/wa_counties.rds)
- [Data-prep script](scripts/build_wa_vignette_data.R)
:::

## Model

Let $i$ index an observed county-year direct estimate, with county $a_i$ and year $t_i$. The direct estimate $\hat{y}_i$ is modeled as

$$
\hat{y}_i = \theta_i + e_i,\qquad e_i \sim N(0,\tau_i^2),
$$

where $\theta_i$ is the latent county-year biomass mean and $\tau_i^2$ is the observation variance for the direct estimate. The latent mean model is

$$
\theta_i = \beta_0 + w_{a_i,t_i},
$$

where $w_{a,t}$ is a separable CAR-time process over counties and years. Its spatial component is defined by the county graph, and its temporal component is defined over the annual support.

The supplied direct-estimate variance $\hat{v}_i$ enters through a scaled residual variance model. With effective sample size $n_i$, the observation variance is

$$
\log(\tau_i^2) =
\log(\kappa) +
\omega_i\log(\hat{v}_i) +
(1 - \omega_i)\log(\tau_0^2),
\qquad
\omega_i = \frac{n_i}{n_i + c}.
$$

The positive scaling parameter $\kappa$ lets the model learn whether the supplied direct-estimate variances are too small or too large overall. The common scale $\tau_0^2$ gives low-sample-size direct estimates a place to shrink toward on the log-variance scale.

Some multi-plot county-years have zero raw direct-estimate variance because all sampled plots have zero biomass. Those rows contain useful information, but a zero observation variance would treat the direct estimate as exact. The data bundle therefore assigns them a small positive floor, equal to the 5th percentile of the positive raw direct-estimate variances. In this article the floor is only a variance template: the scaled residual variance model recalibrates it using $\kappa$, $\tau_0^2$, and $n_i$.

## Data

The data chunk below reads the county polygons and county-year direct estimates. Rows with missing `direct_biomass_model` are retained because they define county-year latent support for recovery and fitted values, but they do not contribute to the likelihood.

```{r read-data}
#| code-summary: "Show data-reading code"
wa_counties <- read_rds(file.path(data_dir, "wa_counties.rds"))
direct_estimates <- read_csv(file.path(data_dir, "wa_direct_estimates.csv"), show_col_types = FALSE)

wa_counties <- wa_counties |>
  mutate(county_fips = as.character(county_fips))

direct_estimates <- direct_estimates |>
  mutate(
    county_fips = as.character(county_fips),
    year = as.integer(year)
  ) |>
  arrange(county_fips, year)

observed_direct <- direct_estimates |>
  filter(!is.na(direct_biomass_model))
```

The table summarizes the area-response data used in this article. County-years with a single FIA plot remain in the county-year support, but they are not used as direct-estimate likelihood rows because their design-based variance cannot be estimated from one plot.

```{r data-summary}
#| code-summary: "Show data summary code"
direct_summary <- tibble(
  quantity = c(
    "counties",
    "years",
    "county-years",
    "model-ready direct estimates",
    "missing model responses",
    "single-plot rows retained but excluded",
    "variance-floor rows",
    "median plot count among modeled county-years",
    "median direct-estimate SE"
  ),
  value = c(
    n_distinct(direct_estimates$county_fips),
    n_distinct(direct_estimates$year),
    nrow(direct_estimates),
    sum(direct_estimates$direct_estimate_in_model),
    sum(is.na(direct_estimates$direct_biomass_model)),
    sum(direct_estimates$direct_estimate_status == "single_plot_no_variance"),
    sum(direct_estimates$direct_estimate_in_model &
          direct_estimates$direct_biomass_vhat_source == "floor", na.rm = TRUE),
    median(observed_direct$n, na.rm = TRUE),
    median(observed_direct$direct_biomass_se_model, na.rm = TRUE)
  )
) |>
  mutate(value = fmt_num(as.numeric(value), 1))

show_table(direct_summary, caption = "County-year direct-estimate data used in the scaled-variance CAR-time model.")
```

The next table shows selected-year data coverage. The model includes all counties in every year, but years with fewer observed direct estimates rely more heavily on the model structure.

```{r year-coverage}
#| code-summary: "Show year coverage code"
year_coverage <- direct_estimates |>
  group_by(year) |>
  summarise(
    counties_with_direct_estimates = sum(direct_estimate_in_model),
    median_plot_n = median(n[n > 0], na.rm = TRUE),
    median_direct_biomass = median(direct_biomass_model, na.rm = TRUE),
    median_direct_se = median(direct_biomass_se_model, na.rm = TRUE),
    .groups = "drop"
  )

show_table(
  year_coverage |>
    filter(year %in% c(1999, 2005, 2010, 2015, 2020, 2024, 2025)) |>
    mutate(
      median_plot_n = fmt_num(median_plot_n, 0),
      median_direct_biomass = fmt_num(median_direct_biomass, 1),
      median_direct_se = fmt_num(median_direct_se, 1)
    ),
  caption = "Selected-year direct-estimate coverage and precision."
)
```

## County Graph

The county graph defines spatial neighbors for the CAR component. Washington has island components, so the graph construction explicitly adds nearest-neighbor bridge edges. The island rule is part of the model definition.

```{r county-graph}
#| code-fold: false
g <- car_graph(
  wa_counties,
  id = "county_fips",
  island = "nearest",
  island_k = 4
)

g$island_added_edges
```

The figure checks the graph visually. Thin lines connect neighboring county centroids; thicker colored lines show bridge edges added for island components.

```{r county-graph-map, fig.width=7, fig.height=5.2}
#| code-summary: "Show graph plotting code"
coord_dat <- data.frame(
  county_fips = wa_counties$county_fips,
  sf::st_coordinates(sf::st_point_on_surface(sf::st_geometry(wa_counties)))
)

edge_index <- which(as.matrix(g$adjacency) != 0, arr.ind = TRUE)
edge_index <- edge_index[edge_index[, "row"] < edge_index[, "col"], , drop = FALSE]
edge_dat <- tibble(
  from = g$ids[edge_index[, "row"]],
  to = g$ids[edge_index[, "col"]]
) |>
  mutate(
    key = paste(pmin(from, to), pmax(from, to), sep = "--"),
    x = coord_dat$X[match(from, coord_dat$county_fips)],
    y = coord_dat$Y[match(from, coord_dat$county_fips)],
    xend = coord_dat$X[match(to, coord_dat$county_fips)],
    yend = coord_dat$Y[match(to, coord_dat$county_fips)]
  )

island_key <- paste(
  pmin(g$island_added_edges$from, g$island_added_edges$to),
  pmax(g$island_added_edges$from, g$island_added_edges$to),
  sep = "--"
)
edge_dat$island <- edge_dat$key %in% island_key

ggplot(wa_counties) +
  geom_sf(fill = "grey96", color = "white", linewidth = 0.15) +
  geom_segment(
    data = edge_dat,
    aes(x = x, y = y, xend = xend, yend = yend, color = island, linewidth = island)
  ) +
  geom_point(data = coord_dat, aes(X, Y), color = stlmm_color("primary"), size = 1.3) +
  coord_sf(expand = FALSE) +
  scale_color_manual(
    values = c("FALSE" = "grey45", "TRUE" = stlmm_color("secondary")),
    guide = "none"
  ) +
  scale_linewidth_manual(values = c("FALSE" = 0.25, "TRUE" = 1), guide = "none")
```

## Fit

The sampler settings below are chosen to give stable posterior summaries for the example while keeping the workflow practical to rerun. For a final analysis, increase the run length as needed and check convergence diagnostics before interpreting results.

```{r sampler-settings}
#| code-fold: false
n_samples <- 12000
burnin <- 6000
chains <- 3
n_keep <- 100
posterior_thin <- max(1L, floor((n_samples - burnin) / n_keep))
posterior_sub_sample <- list(start = burnin + 1, thin = posterior_thin)
chain_control <- list(seed = 1, dispersion = 1.5)
warmup_control <- list(batch_length = 25, min_batches = 10)
summary_parameters <- c(
  "(Intercept)",
  "kappa",
  "tau0_sq",
  "car_time_1_sigma_sq",
  "car_time_1_rho",
  "car_time_1_phi"
)
```

The `n_keep` value controls how many post-burn-in draws per chain are retained for recovery and fitted-value summaries. The fit summary uses all post-burn-in MCMC samples.

```{r cache-settings}
#| include: false
#| purl: false
fit_cache_tag <- paste(
  "exclude-no-raw-var-neff-mcmc",
  paste(n_samples, burnin, chains, sep = "-"),
  sep = "-"
)

posterior_cache_tag <- paste(
  fit_cache_tag,
  paste0("keep-", n_keep),
  sep = "-"
)
```

In the model call below, `direct_biomass_model` is $\hat{y}_i$, `direct_biomass_vhat_model` is $\hat{v}_i$, and `n_eff_model` is $n_i$. Scaled direct-estimate variances enter through `resid(model = "scaled", variance = direct_biomass_vhat_model, n = n_eff_model)`.

```{r fit-model}
#| code-fold: false
#| eval: false
#| purl: true
fit <- stLMM(
  direct_biomass_model ~
    car_time(county_fips, year, graph = g, car_model = "leroux") +
    resid(
      model = "scaled",
      variance = direct_biomass_vhat_model,
      n = n_eff_model,
      shrinkage = 10,
      kappa_log_prior = c(mean = 0, sd = 1)
    ),
  data = direct_estimates,
  priors = list(
    car_time_1 = list(
      sigma_sq = half_t(
        df = 3,
        scale = sd(observed_direct$direct_biomass_model, na.rm = TRUE)
      ),
      rho = uniform(0.01, 0.99),
      phi = uniform(-0.99, 0.99)
    )
  ),
  n_samples = n_samples,
  chains = chains,
  chain_control = chain_control,
  warmup = warmup_control,
  verbose = TRUE,
  n_report = 500
)

fit_summary <- summary(fit, burn = burnin, parameters = summary_parameters)
```

```{r fit-model-execute}
#| include: false
#| purl: false
fit_cache <- use_article_cache(article_id, paste0("fit-", fit_cache_tag), {
  fit <- stLMM(
    direct_biomass_model ~
      car_time(county_fips, year, graph = g, car_model = "leroux") +
      resid(
        model = "scaled",
        variance = direct_biomass_vhat_model,
        n = n_eff_model,
        shrinkage = 10,
        kappa_log_prior = c(mean = 0, sd = 1)
      ),
    data = direct_estimates,
    priors = list(
      car_time_1 = list(
        sigma_sq = half_t(
          df = 3,
          scale = sd(observed_direct$direct_biomass_model, na.rm = TRUE)
        ),
        rho = uniform(0.01, 0.99),
        phi = uniform(-0.99, 0.99)
      )
    ),
    n_samples = n_samples,
    chains = chains,
    chain_control = chain_control,
    warmup = warmup_control,
    verbose = TRUE,
    n_report = 500
  )

  list(
    fit = fit,
    fit_summary = summary(fit, burn = burnin, parameters = summary_parameters)
  )
})

if (is.null(fit_cache$fit_summary)) {
  fit_cache$fit_summary <- summary(fit_cache$fit, burn = burnin, parameters = summary_parameters)
}

list2env(fit_cache, envir = environment())
```

```{r fit-summary}
#| code-summary: "Show fit summary code"
#| purl: true
fit_summary
```

The `kappa` and `tau0_sq` rows in the summary describe how the model recalibrates the direct-estimate variances. The parameter `kappa` is the overall multiplier: values below one shrink the supplied variance template, and values above one inflate it. The parameter `tau0_sq` is the common variance scale used when a county-year has limited FIA support.

The sample-size weight $\omega_i = n_i / (n_i + 10)$ is deterministic. It controls how much each county-year leans on its own supplied direct-estimate variance rather than the common scale. A county-year with $n_i = 10$ gets weight 0.50 on its supplied variance, while one with $n_i = 50$ gets weight 0.83. The practical effect is that low-sample-size direct estimates are not allowed to dictate their own observation variance as strongly as high-sample-size direct estimates.

The scatter plot below compares each supplied direct-estimate variance with the posterior mean of its model-implied observation variance under the scaled model. Points on the 1:1 line would use the supplied variance unchanged. Points below the line have been scaled downward; points above the line have been scaled upward or pulled toward the common variance scale.

```{r scaled-variance-scatter, fig.width=6.5, fig.height=5.5}
#| code-summary: "Show scaled-variance comparison code"
variance_draws <- as_samples(fit, burn = burnin, metadata = FALSE)

scaled_variance_summary <- observed_direct |>
  mutate(
    variance_weight = n_eff_model / (n_eff_model + 10),
    tau_sq_mean = purrr::map2_dbl(
      direct_biomass_vhat_model,
      variance_weight,
      \(vhat_i, weight_i) {
        mean(
          variance_draws$kappa *
            exp(
              weight_i * log(vhat_i) +
                (1 - weight_i) * log(variance_draws$tau0_sq)
            )
        )
      }
    )
  )

variance_axis_limits <- range(
  c(
    scaled_variance_summary$direct_biomass_vhat_model,
    scaled_variance_summary$tau_sq_mean
  ),
  na.rm = TRUE
)
variance_axis_limits <- 10^c(
  floor(log10(variance_axis_limits[1])),
  ceiling(log10(variance_axis_limits[2]))
)

ggplot(
  scaled_variance_summary,
  aes(x = direct_biomass_vhat_model, y = tau_sq_mean)
) +
  geom_abline(slope = 1, intercept = 0, color = "grey45", linewidth = 0.45) +
  geom_point(aes(color = n_eff_model), alpha = 0.65, size = 1.7) +
  scale_x_log10(
    limits = variance_axis_limits,
    labels = scales::label_number(big.mark = ",", trim = TRUE)
  ) +
  scale_y_log10(
    limits = variance_axis_limits,
    labels = scales::label_number(big.mark = ",", trim = TRUE)
  ) +
  scale_color_viridis_c(option = "C", name = "Plot n") +
  coord_fixed() +
  labs(
    x = "Supplied direct-estimate variance",
    y = "Posterior mean scaled observation variance"
  )
```

## Recovery and Fitted Means

Structured process terms are collapsed during Gaussian model fitting. The `recover()` call draws the county-year CAR-time effects from retained posterior parameter draws so fitted values can include the latent process contribution.

```{r recover-model}
#| code-fold: false
#| eval: false
#| purl: true
rec <- recover(
  fit,
  sub_sample = posterior_sub_sample
)
```

```{r recover-model-execute}
#| include: false
#| purl: false
recovery_cache <- use_article_cache(article_id, paste0("recovery-", posterior_cache_tag), {
  rec <- recover(
    fit,
    sub_sample = posterior_sub_sample
  )

  list(rec = rec)
})

list2env(recovery_cache, envir = environment())
```

The fitted values are posterior draws of the latent county-year mean for the same county-year support used in the fit. This support includes county-years with missing direct estimates. Those rows did not contribute to the likelihood, but their CAR-time effects can still be recovered because their county and year define latent support nodes.

```{r fitted-county-years}
#| code-summary: "Show fitted-value summary code"
#| code-fold: true
#| eval: false
#| purl: true
fitted_draws <- as.matrix(as_samples(fitted(rec, summary = FALSE), metadata = FALSE))

county_year_summary <- direct_estimates |>
  bind_cols(
    summarize_draw_matrix(fitted_draws, prefix = "theta_") |>
      select(-prediction_row)
  )
```

```{r fitted-county-years-execute}
#| include: false
#| purl: false
fitted_cache <- use_article_cache(article_id, paste0("fitted-", posterior_cache_tag), {
  fitted_draws <- as.matrix(as_samples(fitted(rec, summary = FALSE), metadata = FALSE))

  county_year_summary <- direct_estimates |>
    bind_cols(
      summarize_draw_matrix(fitted_draws, prefix = "theta_") |>
        select(-prediction_row)
    )

  list(
    fitted_draws = fitted_draws,
    county_year_summary = county_year_summary
  )
})

list2env(fitted_cache, envir = environment())
```

The resulting summary table has the common output shape used throughout the series: one row per county-year, direct-estimate information where available, and posterior means and 95% credible intervals for the model-based county-year biomass mean.

```{r selected-summary-table}
#| code-summary: "Show selected county-year summary code"
show_table(
  county_year_summary |>
    filter(year == 2024) |>
    arrange(desc(theta_mean)) |>
    select(
      county, n, direct_biomass_model, direct_biomass_se_model,
      theta_mean, theta_lower, theta_upper
    ) |>
    slice_head(n = 10) |>
    mutate(
      across(
        c(direct_biomass_model, direct_biomass_se_model, theta_mean, theta_lower, theta_upper),
        ~ fmt_num(.x, 1)
      )
    ),
  caption = "Highest 2024 model-smoothed county-year biomass means, with 95% posterior credible intervals."
)
```

## Direct Estimates and Smoothed Means

The next figure compares observed direct estimates with model-smoothed county-year means. Counties with no direct estimate in a year can still receive a model-based estimate because their county-year latent effect is part of the CAR-time support.

```{r direct-vs-smoothed-map, fig.width=10, fig.height=3.8}
#| code-summary: "Show direct-versus-smoothed map code"
panel_years <- c(2005, 2010, 2015, 2020, 2024)

map_dat <- wa_counties |>
  left_join(
    county_year_summary |> filter(year %in% panel_years),
    by = "county_fips"
  )

map_long <- bind_rows(
  map_dat |> mutate(quantity = "direct estimate", value = direct_biomass_model),
  map_dat |> mutate(quantity = "model-smoothed\nmean", value = theta_mean)
) |>
  mutate(quantity = factor(quantity, levels = c("direct estimate", "model-smoothed\nmean")))

ggplot(map_long) +
  geom_sf(aes(fill = value), color = "grey80", linewidth = 0.12) +
  coord_sf(expand = FALSE) +
  facet_grid(quantity ~ year) +
  scale_fill_gradientn(
    colors = stlmm_palette(),
    name = "Mg/ha",
    na.value = "grey92"
  ) +
  theme(
    panel.spacing = grid::unit(0.02, "lines"),
    strip.text = element_text(margin = margin(1, 1, 1, 1)),
    plot.margin = margin(0, 0, 0, 0)
  )
```

The scatter plot shows the same comparison only for county-years used in the direct-estimate likelihood. The model is not trying to reproduce every noisy direct estimate; it smooths them according to the scaled variance model and the county-time dependence structure.

```{r direct-vs-smoothed-scatter, fig.width=6.5, fig.height=5.5}
#| code-summary: "Show direct-versus-smoothed scatter code"
scatter_dat <- county_year_summary |>
  filter(!is.na(direct_biomass_model))

fit_axis_limits <- range(
  c(scatter_dat$direct_biomass_model, scatter_dat$theta_mean, 0),
  na.rm = TRUE
)

ggplot(scatter_dat, aes(x = direct_biomass_model, y = theta_mean)) +
  geom_abline(slope = 1, intercept = 0, color = "grey45", linewidth = 0.45) +
  geom_point(
    aes(size = 1 / direct_biomass_vhat_model, color = year),
    alpha = 0.55
  ) +
  scale_size_continuous(range = c(0.5, 3.5), guide = "none") +
  scale_color_viridis_c(option = "C") +
  coord_equal(xlim = fit_axis_limits, ylim = fit_axis_limits) +
  labs(
    x = "Direct estimate Mg/ha",
    y = "CAR-time posterior mean Mg/ha",
    color = "Year"
  )
```

## County Time Series

Time series show how the model smooths through years with sparse or missing direct estimates. The example counties below are selected to include different FIA units and biomass ranges.

```{r county-time-series, fig.width=10, fig.height=6.5}
#| code-summary: "Show county time-series code"
selected_counties <- c("Clallam", "King", "Okanogan", "Yakima")

series_dat <- county_year_summary |>
  filter(county %in% selected_counties) |>
  mutate(county = factor(county, levels = selected_counties))

ggplot(series_dat, aes(x = year)) +
  geom_ribbon(
    aes(ymin = theta_lower, ymax = theta_upper),
    fill = "#9ecae1",
    alpha = 0.45
  ) +
  geom_line(aes(y = theta_mean), color = stlmm_color("primary"), linewidth = 0.7) +
  geom_point(
    aes(y = direct_biomass_model, size = n),
    color = stlmm_color("secondary"),
    alpha = 0.75,
    na.rm = TRUE
  ) +
  facet_wrap(~ county, ncol = 2, scales = "free_y") +
  scale_x_continuous(breaks = seq(min(series_dat$year), max(series_dat$year), by = 5)) +
  scale_size_continuous(range = c(1, 3.5), name = "Plot n") +
  labs(
    x = "Year",
    y = "Live aboveground biomass Mg/ha"
  )
```

The ribbons are 95% posterior credible intervals for the latent county-year mean. Points are model-ready direct estimates, with point size proportional to the number of FIA plot rows used in that county-year direct estimate. These intervals do not include posterior predictive error for a new direct estimate; the direct-estimate observation variance is represented by the noisy points, not added to the ribbon.

## Scaled Direct-Estimate Variances

The fixed-variance Fay-Herriot formulation treats the supplied direct-estimate variances as known. That is a useful baseline, but it can be uncomfortable in forest inventory applications because the direct-estimate variances are themselves estimated from the sample.

The scaled residual variance model is a practical relaxation. The supplied direct-estimate variances still describe relative precision across county-years, but the model estimates how strongly those variances should be trusted. With `n_eff_model`, lower-sample-size direct estimates shrink more toward a common variance scale, while higher-sample-size direct estimates stay closer to their supplied variance.

Another option is to estimate one observation variance per direct estimate with `resid(model = "group", variance = ..., prior = "ig")`, or to use the effective-sample-size construction behind a Shannon-style prior [@Shannon2025]. Those models can be useful when there are relatively few direct estimates or when the sampler is configured for a higher-dimensional residual variance block. For this Washington county-year dataset, they add hundreds of variance parameters, so the scaled model is the cleaner example.

WAIC can be used to compare fixed- and scaled-variance formulations only when they are fit to the same likelihood rows. The fixed-variance article deliberately excludes the floored zero-variance rows, so a formal WAIC comparison would require a matched sensitivity fit. This article leaves that comparison to the series bake-off so the focus stays on the variance model itself.

## Summary

This article shows the first residual-variance extension to the area-response workflow:

- the response rows are model-ready county-year direct estimates;
- single-plot county-years remain in the county-year support but do not enter the direct-estimate likelihood;
- scaled direct-estimate variances enter through `resid(model = "scaled", variance = ..., n = ...)`;
- the county graph defines spatial borrowing;
- `car_time()` adds separable county-time smoothing;
- `recover()` is needed before computing fitted values with Gaussian structured-process models;
- the common output is a county-year table of posterior means and 95% credible intervals.

The next area-response article adds county mean TCC as a covariate. That lets the model borrow information not only from neighboring counties and adjacent years, but also from remote-sensing variation that is observed for every county-year.
