---
title: "Unit-response two-stage CAR-time model with FIA-unit variances"
subtitle: "A residual-variance refinement for Washington biomass"
author: "Andrew O. Finley"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-overflow: wrap
    html-math-method: katex
    css: stlmm-vignette.css
execute:
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| code-summary: "Show setup code"
library(stLMM)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)

source("article-utils.R")

set.seed(1)
theme_set(theme_bw(base_size = 12))

article_id <- "wa-unit-two-stage-car-time-tcc-fiaunit-var"
data_dir <- "wa_data"
```

## Purpose

This article keeps the same two-stage CAR-time TCC mean model from the previous unit-response article and changes only the positive-biomass residual variance. The binary stage still models `agb_live_mg_ha > 0`; the positive stage still models square-root positive `agb_live_mg_ha`; and the county-year target remains the grid-supported mean of `agb_live_mg_ha`.

The variance refinement is useful because Washington counties span large biomass gradients, from relatively low-biomass dry interior forests to high-biomass coastal and western forests. A common residual variance asks the positive-biomass model to use one plot-level noise scale across that full range. Allowing the residual variance to differ by FIA unit gives the positive stage a simple way to accommodate broad heteroskedasticity while leaving the occurrence model, mean structure, and aggregation target unchanged.

The extension is

$$
q_j = \delta_0 + \delta_1 x_j + w^+_{a_j,t_j} + r_j,\qquad
r_j \sim \operatorname{normal}(0,\tau^2_{u_j}),
$$

where $q_j = \sqrt{y_j^+}$, $x_j$ is standardized row-level TCC, $w^+_{a,t}$ is the positive-stage CAR-time process, and $u_j$ is the FIA unit for plot row $j$. This is a variance-model refinement, not a new estimand or a new mean structure.

::: {.stlmm-resource-box .stlmm-resource-source}
<div class="stlmm-resource-title">Source Code</div>

- [Quarto source](wa-unit-two-stage-car-time-tcc-fiaunit-var.qmd)
- [Extracted R code](wa-unit-two-stage-car-time-tcc-fiaunit-var.R)
:::

::: {.stlmm-resource-box .stlmm-resource-data}
<div class="stlmm-resource-title">Data Bundle</div>

- [Download Washington biomass data](wa_data.zip)
:::

## Data

```{r read-data}
#| code-summary: "Show data-reading code"
wa_counties <- read_rds(file.path(data_dir, "wa_counties.rds"))
unit_plots <- read_csv(file.path(data_dir, "wa_unit_plots.csv"), show_col_types = FALSE)
direct_estimates <- read_csv(file.path(data_dir, "wa_direct_estimates.csv"), show_col_types = FALSE)
prediction_grid <- read_csv(
  file.path(data_dir, "wa_unit_prediction_grid_5km.csv"),
  show_col_types = FALSE
)

wa_counties <- wa_counties |>
  mutate(
    county_fips = as.character(county_fips),
    fia_unit = factor(fia_unit)
  )

unit_plots <- unit_plots |>
  mutate(
    county_fips = as.character(county_fips),
    county_fips = factor(county_fips),
    fia_unit = factor(fia_unit),
    year = as.integer(year),
    biomass_positive = as.integer(agb_live_mg_ha > 0)
  ) |>
  arrange(county_fips, year, plot_id)

direct_estimates <- direct_estimates |>
  mutate(
    county_fips = as.character(county_fips),
    year = as.integer(year)
  )

plot_years <- sort(unique(unit_plots$year))
tcc_center <- mean(unit_plots$tcc_mean, na.rm = TRUE)
tcc_scale <- sd(unit_plots$tcc_mean, na.rm = TRUE)

unit_plots <- unit_plots |>
  mutate(tcc_mean_scaled = (tcc_mean - tcc_center) / tcc_scale)

county_levels <- levels(unit_plots$county_fips)
fia_unit_levels <- levels(unit_plots$fia_unit)
fia_unit_names <- unit_plots |>
  distinct(fia_unit, fia_unit_name) |>
  arrange(fia_unit)

prediction_grid <- prediction_grid |>
  mutate(
    county_fips = as.character(county_fips),
    county_fips = factor(county_fips, levels = county_levels),
    fia_unit = factor(fia_unit, levels = fia_unit_levels),
    year = as.integer(year),
    tcc_mean_scaled = (tcc_mean - tcc_center) / tcc_scale
  ) |>
  filter(year %in% plot_years) |>
  arrange(county_fips, year, x, y)

positive_plots <- unit_plots |>
  filter(biomass_positive == 1) |>
  mutate(biomass_sqrt = sqrt(agb_live_mg_ha))
```

The plot and grid TCC variables are annual point samples from the USDA Forest Service Tree Canopy Cover raster products [@USFSTreeCanopyCover2026]. This article uses the same plot-based TCC standardization and 5 km prediction support as the previous two-stage CAR-time TCC article.

```{r data-summary}
#| code-summary: "Show data summary code"
data_summary <- tibble(
  quantity = c(
    "plot rows",
    "positive-biomass plot rows",
    "FIA units",
    "5 km prediction-grid rows"
  ),
  value = c(
    fmt_int(nrow(unit_plots)),
    fmt_int(nrow(positive_plots)),
    fmt_int(n_distinct(unit_plots$fia_unit)),
    fmt_int(nrow(prediction_grid))
  )
)

show_table(data_summary, caption = "Unit-response data for the FIA-unit variance model.")
```

## County Graph

The CAR-time mean structure is unchanged from the previous article.

```{r county-graph}
#| code-fold: false
g <- car_graph(
  wa_counties,
  id = "county_fips",
  island = "nearest",
  island_k = 4
)

g$island_added_edges
```

## Fit

Only the positive-biomass stage changes. The binary occurrence stage is the same Bernoulli CAR-time TCC model used in the previous article.

```{r mcmc-settings}
#| code-summary: "Show MCMC settings"
n_samples <- 10000
burnin <- 5000
chains <- 3
n_keep <- 100
posterior_thin <- max(1L, floor((n_samples - burnin) / n_keep))
posterior_sub_sample <- list(start = burnin + 1, thin = posterior_thin)
chain_control <- list(seed = 17, dispersion = 1.5)
warmup_control <- list(batch_length = 25, min_batches = 10)

sqrt_response_sd <- sd(positive_plots$biomass_sqrt, na.rm = TRUE)
```

```{r cache-settings}
#| include: false
#| purl: false
fit_cache_tag <- paste(
  "two-stage-car-time-tcc-fiaunit-var-mcmc",
  paste(n_samples, burnin, chains, sep = "-"),
  sep = "-"
)

posterior_cache_tag <- paste(
  fit_cache_tag,
  paste0("keep-", n_keep),
  sep = "-"
)

fit_cache_name <- paste0("fits-", fit_cache_tag)
recovery_cache_name <- paste0("recovery-", posterior_cache_tag)
prediction_cache_name <- paste0("prediction-", posterior_cache_tag)
```

```{r fit-occurrence}
#| code-fold: false
#| eval: false
#| purl: true
occurrence_fit <- stLMM(
  biomass_positive ~
    tcc_mean_scaled +
    car_time(county_fips, year, graph = g, car_model = "leroux"),
  data = unit_plots,
  family = "binomial",
  priors = list(
    beta = normal(mean = 0, sd = 2.5),
    car_time_1 = list(
      sigma_sq = half_t(df = 3, scale = 1),
      rho = uniform(0.01, 0.99),
      phi = uniform(-0.99, 0.99)
    )
  ),
  n_samples = n_samples,
  chains = chains,
  chain_control = chain_control,
  warmup = warmup_control,
  verbose = TRUE,
  n_report = 500
)
```

```{r fit-positive-biomass}
#| code-fold: false
#| eval: false
#| purl: true
biomass_fit <- stLMM(
  biomass_sqrt ~
    tcc_mean_scaled +
    car_time(county_fips, year, graph = g, car_model = "leroux") +
    resid(model = "group", group = fia_unit),
  data = positive_plots,
  priors = list(
    beta = normal(mean = 0, sd = 2 * sqrt_response_sd),
    resid = list(tau_sq = half_t(df = 3, scale = sqrt_response_sd)),
    car_time_1 = list(
      sigma_sq = half_t(df = 3, scale = sqrt_response_sd),
      rho = uniform(0.01, 0.99),
      phi = uniform(-0.99, 0.99)
    )
  ),
  n_samples = n_samples,
  chains = chains,
  chain_control = chain_control,
  warmup = warmup_control,
  verbose = TRUE,
  n_report = 500
)
```

```{r fit-models-execute}
#| include: false
#| purl: false
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
fit_cache <- use_article_cache(article_id, fit_cache_name, {
  occurrence_fit <- stLMM(
    biomass_positive ~
      tcc_mean_scaled +
      car_time(county_fips, year, graph = g, car_model = "leroux"),
    data = unit_plots,
    family = "binomial",
    priors = list(
      beta = normal(mean = 0, sd = 2.5),
      car_time_1 = list(
        sigma_sq = half_t(df = 3, scale = 1),
        rho = uniform(0.01, 0.99),
        phi = uniform(-0.99, 0.99)
      )
    ),
    n_samples = n_samples,
    chains = chains,
    chain_control = chain_control,
    warmup = warmup_control,
    verbose = TRUE,
    n_report = 500
  )

  biomass_fit <- stLMM(
    biomass_sqrt ~
      tcc_mean_scaled +
      car_time(county_fips, year, graph = g, car_model = "leroux") +
      resid(model = "group", group = fia_unit),
    data = positive_plots,
    priors = list(
      beta = normal(mean = 0, sd = 2 * sqrt_response_sd),
      resid = list(tau_sq = half_t(df = 3, scale = sqrt_response_sd)),
      car_time_1 = list(
        sigma_sq = half_t(df = 3, scale = sqrt_response_sd),
        rho = uniform(0.01, 0.99),
        phi = uniform(-0.99, 0.99)
      )
    ),
    n_samples = n_samples,
    chains = chains,
    chain_control = chain_control,
    warmup = warmup_control,
    verbose = TRUE,
    n_report = 500
  )

  list(
    occurrence_fit = occurrence_fit,
    occurrence_summary = summary(occurrence_fit, burn = burnin),
    biomass_fit = biomass_fit,
    biomass_summary = summary(biomass_fit, burn = burnin)
  )
})

list2env(fit_cache, envir = environment())

prefix_group_residual_variance_names <- function(fit) {
  if (inherits(fit, "stLMM_chains")) {
    fit$chains <- lapply(fit$chains, prefix_group_residual_variance_names)
    return(fit)
  }

  if (!is.null(fit$residual_variance_samples) &&
      is.matrix(fit$residual_variance_samples)) {
    nm <- colnames(fit$residual_variance_samples)
    if (!is.null(nm) && !all(startsWith(nm, "tau_sq_"))) {
      colnames(fit$residual_variance_samples) <- paste0("tau_sq_", nm)
      fit$samples$residual_variance <- fit$residual_variance_samples
    }
  }

  fit
}

biomass_fit <- prefix_group_residual_variance_names(biomass_fit)
biomass_summary <- summary(biomass_fit, burn = burnin)
```

```{r fit-summaries}
#| code-summary: "Show fit summaries"
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
occurrence_summary
biomass_summary
```

The positive-stage summary now includes one residual variance parameter for each FIA unit. The map below shows the corresponding residual standard deviations on the square-root biomass scale.

```{r fiaunit-residual-sd-map, fig.width=8, fig.height=4.8}
#| code-summary: "Show FIA-unit residual SD map code"
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
biomass_draws <- as_samples(biomass_fit, burn = burnin, metadata = FALSE)
resid_var_names <- paste0("tau_sq_", fia_unit_levels)
if (!all(resid_var_names %in% colnames(biomass_draws))) {
  resid_var_names <- fia_unit_levels
}
resid_var_draws <- biomass_draws[, resid_var_names, drop = FALSE]
resid_sd_draws <- sqrt(resid_var_draws)

fia_resid_sd <- tibble(
  fia_unit = factor(fia_unit_levels, levels = fia_unit_levels),
  sd_mean = colMeans(resid_sd_draws),
  sd_lower = apply(resid_sd_draws, 2, quantile, probs = 0.025),
  sd_upper = apply(resid_sd_draws, 2, quantile, probs = 0.975)
) |>
  left_join(fia_unit_names, by = "fia_unit")

fia_resid_map <- wa_counties |>
  select(-fia_unit_name) |>
  left_join(fia_resid_sd, by = "fia_unit")

fia_resid_labels <- fia_resid_map |>
  group_by(fia_unit, fia_unit_name, sd_mean) |>
  summarise(geometry = sf::st_union(geometry), .groups = "drop") |>
  mutate(label = paste0(fia_unit_name, "\n", fmt_num(sd_mean, 2))) |>
  sf::st_point_on_surface()

ggplot(fia_resid_map) +
  geom_sf(aes(fill = sd_mean), color = "grey80", linewidth = 0.15) +
  geom_sf_label(
    data = fia_resid_labels,
    aes(label = label),
    size = 3,
    label.size = 0.12,
    fill = "white",
    alpha = 0.85
  ) +
  coord_sf(expand = FALSE) +
  scale_fill_gradientn(
    colors = stlmm_palette(),
    name = "resid. SD",
    na.value = "grey92"
  ) +
  labs(
    title = "positive-stage residual SD by FIA unit",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(size = 10, face = "bold", margin = margin(0, 0, 2, 0)),
    plot.margin = margin(0, 0, 0, 0)
  )
```

```{r fiaunit-residual-sd-table}
#| code-summary: "Show FIA-unit residual SD table code"
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
show_table(
  fia_resid_sd |>
    transmute(
      `FIA unit` = fia_unit_name,
      `posterior mean SD` = fmt_num(sd_mean, 2),
      `lower` = fmt_num(sd_lower, 2),
      `upper` = fmt_num(sd_upper, 2)
    ),
  caption = "FIA-unit residual standard deviations for square-root positive biomass."
)
```

## Prediction and Aggregation

Prediction is unchanged: draw-wise occurrence predictions are multiplied by squared positive-biomass predictions, then averaged within county-year. The Gaussian positive-stage predictions use `y_samples = TRUE`, so the squared positive-biomass draws include residual predictive uncertainty before recombination.

```{r aggregation-helpers}
#| code-summary: "Show aggregation helper code"
prediction_sample_matrix <- function(pred, sample) {
  as.matrix(as_samples(pred, sample = sample, metadata = FALSE))
}

aggregate_grid_draws <- function(draws, grid, prefix = "") {
  county_year <- grid |>
    distinct(county_fips, county, year) |>
    arrange(county_fips, year)

  county_samples <- matrix(
    NA_real_,
    nrow = nrow(draws),
    ncol = nrow(county_year)
  )

  for (j in seq_len(nrow(county_year))) {
    ii <- which(
      grid$county_fips == county_year$county_fips[j] &
        grid$year == county_year$year[j]
    )
    county_samples[, j] <- rowMeans(draws[, ii, drop = FALSE])
  }

  county_year |>
    bind_cols(
      summarize_draw_matrix(county_samples, prefix = prefix) |>
        select(-prediction_row)
    )
}
```

```{r recover-models}
#| code-fold: false
#| eval: false
#| purl: true
occurrence_rec <- recover(
  occurrence_fit,
  sub_sample = posterior_sub_sample
)

biomass_rec <- recover(
  biomass_fit,
  sub_sample = posterior_sub_sample
)
```

```{r recover-models-execute}
#| include: false
#| purl: false
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
recovery_cache <- use_article_cache(article_id, recovery_cache_name, {
  occurrence_rec <- recover(
    occurrence_fit,
    sub_sample = posterior_sub_sample
  )

  biomass_rec <- recover(
    biomass_fit,
    sub_sample = posterior_sub_sample
  )

  list(
    occurrence_rec = occurrence_rec,
    biomass_rec = biomass_rec
  )
})

list2env(recovery_cache, envir = environment())
```

```{r predict-grid}
#| code-fold: false
#| eval: false
#| purl: true
occurrence_pred <- predict(
  occurrence_rec,
  newdata = prediction_grid,
  y_samples = TRUE
)

biomass_pred <- predict(
  biomass_rec,
  newdata = prediction_grid,
  y_samples = TRUE
)
```

```{r predict-grid-execute}
#| include: false
#| purl: false
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
prediction_cache <- use_article_cache(article_id, prediction_cache_name, {
  occurrence_pred <- predict(
    occurrence_rec,
    newdata = prediction_grid,
    y_samples = TRUE
  )

  biomass_pred <- predict(
    biomass_rec,
    newdata = prediction_grid,
    y_samples = TRUE
  )

  occurrence_draws <- prediction_sample_matrix(occurrence_pred, "y")
  biomass_sqrt_draws <- prediction_sample_matrix(biomass_pred, "y")

  n_draw <- min(nrow(occurrence_draws), nrow(biomass_sqrt_draws))
  response_draws <- occurrence_draws[seq_len(n_draw), , drop = FALSE] *
    biomass_sqrt_draws[seq_len(n_draw), , drop = FALSE]^2

  county_summary <- aggregate_grid_draws(
    response_draws,
    prediction_grid,
    prefix = "theta_"
  )

  rm(occurrence_pred, biomass_pred, occurrence_draws, biomass_sqrt_draws, response_draws)

  list(county_summary = county_summary)
})

list2env(prediction_cache, envir = environment())
```

```{r combine-direct-estimates}
#| code-summary: "Show direct-estimate join code"
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
direct_for_compare <- direct_estimates |>
  filter(year %in% plot_years) |>
  select(
    county_fips, county, year, n,
    direct_biomass_model, direct_biomass_se_model
  )

county_summary <- county_summary |>
  mutate(county_fips = as.character(county_fips)) |>
  left_join(
    direct_for_compare,
    by = c("county_fips", "county", "year")
  )
```

```{r selected-summary-table}
#| code-summary: "Show selected county-year summary code"
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
show_table(
  county_summary |>
    filter(year == 2024) |>
    arrange(desc(theta_mean)) |>
    select(county, n, direct_biomass_model, theta_mean, theta_lower, theta_upper) |>
    slice_head(n = 10) |>
    mutate(
      across(
        c(direct_biomass_model, theta_mean, theta_lower, theta_upper),
        ~ fmt_num(.x, 1)
      )
    ),
  caption = "Highest 2024 county-year posterior predictive means from the FIA-unit variance model."
)
```

```{r county-time-series, fig.width=9, fig.height=5.8}
#| code-summary: "Show county time-series code"
#| eval: !expr identical(tolower(Sys.getenv("STLMM_RUN_LONG_ARTICLES", unset = "false")), "true")
profile_counties <- c("Clallam", "King", "Okanogan", "Yakima")

profile_dat <- county_summary |>
  filter(county %in% profile_counties) |>
  mutate(county = factor(county, levels = profile_counties))

ggplot(profile_dat, aes(year, theta_mean)) +
  geom_ribbon(
    aes(ymin = theta_lower, ymax = theta_upper),
    fill = stlmm_color("primary"),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(color = stlmm_color("primary"), linewidth = 0.75) +
  geom_point(
    aes(y = direct_biomass_model),
    color = "grey20",
    size = 1.4,
    alpha = 0.75,
    na.rm = TRUE
  ) +
  facet_wrap(~ county, scales = "free_y", ncol = 2) +
  labs(
    x = "year",
    y = "mean agb_live_mg_ha (Mg/ha)"
  )
```

Compared with the previous common-variance model, Okanogan is shifted downward in this figure. Okanogan belongs to the Central FIA unit, which has a smaller estimated positive-stage residual variance than the pooled residual variance used in the previous article. Because the positive stage is fit on the square-root scale and predictions are squared to return to Mg/ha, this residual variance contributes to the back-transformed positive-biomass mean. A smaller residual variance therefore lowers the recombined county-year mean even when the fitted mean structure changes little.

The direct estimates remain a point of comparison, not the model target. This pattern shows what the variance refinement is doing to the model-based grid summary; it does not by itself imply that the grouped-variance model is worse because it is farther from some direct estimates.

## Takeaways

This article changes the residual variance model, not the target or the mean structure. FIA-unit-specific residual variances let the positive Gaussian stage represent different amounts of plot-level spread across broad FIA units while keeping the same occurrence model, TCC effect, and county-time process.

The main diagnostic is whether the grouped residual SDs differ enough to matter and whether the county-year predictions remain consistent with the previous two-stage CAR-time TCC article.
